Fitting bifurcation structure, not voltage traces: A biophysically inspired derivation of reduced neuron models exemplified by potassium dynamics

Version 1

Abstract

Abstract

In conductance-based models, spiking-induced ion concentrations fluctuations can modify single neurons’ excitability. What are the consequences in networks? To study this, simple models capturing ion concentration dynamics realistically are needed. We propose a method to derive a phenomenological model capturing the coupled extracellular potassium and voltage dynamics from a given class I conductance-based model. Rather than fitting voltage traces, we fit the bifurcation structure of the target model, thereby capturing parameter heterogeneity and rich dynamics. The resulting model extends the quadratic integrate-and-fire model, with extracellular potassium accumulation altering voltage dynamics by increasing the reset voltage. We apply our systematic reduction procedure to the Wang-Buzsáki model. Its phenomenological version exhibits quantitatively comparable dynamics and replicates the reshaping of the phase-response curve associated with the transition from SNIC to HOM spikes at elevated potassium. To illustrate the derived model’s applicability, we explore how changes in potassium concentration influence synchronization in networks.

Introduction

The biophysical properties of neurons are well-established as critical determinants of both single-cell computation and network behavior. While complex neuron models allow a detailed analysis at the single-cell level, investigation of network dynamics often requires simpler models that can retain essential features of neuronal computation yet ensure computational efficiency [1–3]. This challenge is further intensified when biophysical single-cell properties change dynamically, such as through alterations in the local environment (e.g., temperature fluctuations, changing ionic concentrations) or neuromodulatory effects. To better understand how such modifications influence network dynamics, it is essential to also incorporate these physiological changes into simplified neuron models. Here, we derive phenomenological neuron models of the integrate-and-fire type that quantitatively capture the dynamics of more complex conductance-based models not only for one specific set of the model’s parameters but also when a physiologically relevant parameter is allowed to vary.

In contrast to models fitted to experimentally measured voltage traces at essentially one specific point in biophysical parameter space, the simplified model captures the two-dimensional bi-furcation structure spanned by the input current as well as an additional physiologically relevant parameter of interest. Moreover, our specific interest with the approach taken is to adequately represent the neuronal response not only in terms of the firing rate but – because of its computational relevance – also the bifurcation that underlies the onset of regular firing at threshold, including switches in bifurcation type when the physiological parameter varies. While we chose to develop phenomenological models capturing the influence of extracellular potassium in this study, the approach is equally amenable to many other biological parameters.

In the processing of information within single cells, the dynamics of action-potential generation play an important role. Although influenced by numerous cellular characteristics, qualitative differences between regular action-potential firing (with all-or-none pulses at threshold) can only arise from three distinct mathematical mechanisms. Their classification coincides but is not entirely identical with the excitability classes 1 and 2 described by Hodgkin (1948) [4] and observed earlier by Arvanitaki (1938) [5], which are based on the frequency of firing at spike onset (with arbitrarily low frequencies for class I and an immediate jump to higher frequencies for class II). These two classes are associated with distinct mathematical bifurcations that underlie the onset of regular spiking: class I firing corresponding to a saddle-node-on-invariant circle (SNIC) bifurcation, while class II firing typically arises from a subcritical Andronov-Hopf bifurcation. A third bifurcation, the saddle-homoclinic orbit (HOM) bifurcation, can also occur and combines properties of class I and class II behaviors, depending on the mode of analysis. The type of bifurcation is a critical determinant of the network state. For instance, neurons with SNIC dynamics tend to produce asynchronous activity when weakly coupled within inhibitory networks. In contrast, the transition to HOM dynamics in the same network configuration results in synchronous activity.

The transition from SNIC to HOM is of particular interest, as it is ubiquitous in the underlying mathematical framework (despite a quantitative dependence of the switching point on other parameters) and has the potential to induce dramatic changes in network behavior. One well-established parameter that induces this transition is the concentration of extracellular potassium. It is directly affected by neuronal spiking. While the outflow of K+ ions at each spike is typically compensated for by mechanisms such as Na+/K+-ATPases and glial buffering, these may not suffice to prevent transient changes during intense activity [6]. Such elevations can alter the neuron’s excitability, forming a feedback loop that can create rich dynamics [7]. Moreover, abnormally elevated levels of extracellular potassium are involved in paroxysmal pathological activities that are specific to epilepsy (seizures) and migraine with aura (spreading depolarization) [8].

To include the effect of extracellular potassium concentration in simpler, network-compatible phenomenological models, we start from a conductance-based model including the potassium concentration dynamics and fit a simpler, integrate-and-fire-type model equipped with a potassium variable to the conductance based model, aiming to capture both firing rates and the two-dimensional bifurcation structure with input current and extracellular potassium concentration as parametric dimensions. Fitting the bifurcation structure of the reduced model near the SNL bifurcation accurately captures the conductance-based model behavior under a range of ionic concentrations, associated with qualitative changes in excitability. This permits the study of transitions between those different regimes as potassium evolves dynamically. Accordingly, the phenomenological model captures a transition from class I (SNIC) to HOM firing with increases in extracellular potassium. The approach contrasts the more widespread practice of deriving phenomenological models via a fit of spike trains or voltage traces from electrophysiological recordings, as they have been used in international competitions, see for example [9, 10].

In this paper, we first demonstrate how to quantitatively derive the integrate-and-fire type version of a given class I conductance-based model. We then compare, at the single-neuron level, the dynamics of the two models, and their phase-response curves. Finally, we investigate the effect of potassium on neuronal synchronization at the network level.

Results

Reducing a class I conductance-based model to a QIF neuron with dynamic

We demonstrate our procedure on a Wang-Buzsáki model embedded in an extracellular space with dynamic potassium concentration (Fig. 1a) [7, 11]. We reduce it to a model of type quadratic integrate-and-fire (QIF), which allows us to capture the normal forms of both SNIC and HOM bifurcations, as well as the transition between them at the so-called saddle-node-loop (SNL) point via the reset voltage. Reset values below the voltage at the saddle-node bifurcation of the QIF correspond to SNIC onsets, while larger values correspond to HOM onsets. We carefully choose parameter values of the QIF system and incorporate potassium dependencies to match the homoclinic, saddle-node and SNIC bifurcations branches of the target model near the SNL transition. In particular, potassium concentration feeds back on the reset voltage, inducing the SNL. An additional reset rule, which increments potassium concentration by a small amount at each spike, mimics the flow of ions through voltage-gated potassium channels (Fig. 1b). Including an electrogenic Na+/K+-ATPase complements the potassium dynamics. The resulting reduced model is, like the original, a slow-fast system where slow changes in potassium drive the neuron across different excitability regimes.

Schematic representation of the two models: the target model (a) and its corresponding phenomenological version (b).

(a) Conductance-based model of excitability class I with dynamic extracellular potassium concentration, equipped with the usual Na+ and K+ voltage-gated channels and with an electrogenic Na+/K+ ATPase (diagram reproduced from [7]). (b) Quadratic integrate-and-fire (QIF) model with dynamic extracellular potassium concentration, equipped with an electrogenic Na+/K+ ATPase. Each spike (i.e., when the voltage v exceeds a threshold vth) triggers two reset rules: as in the classical QIF model the voltage is reset to a value vr, and the extracellular potassium concentration is additionally incremented by a small amount ΔK.

Original model

For convenience, we reproduce in “Methods” the definition of the extended Wang-Buzsáki model [7], that we use as example to illustrate the reduction procedure.

Relying on the difference in timescale between and the other variables, the fast subsystem of System (4) can defined as Eqs. (4a) to (4c), where is a parameter. At physiological potassium concentrations spike onset is mediated by a SNIC bifurcation (e.g. Fig. 2A when ), while a larger concentrations it is mediated by a homoclinic bifurcation (Fig. 2B, ) [7]. Fig. 3A shows the SNL bifurcation organizing this transition.

Matching the bifurcation structure of the target model with the QIF model, at fixed values of .

(a)-(b) Bifurcation diagrams of the fast subsystem of the Wang-Buzsáki model (Eqs. (4a) to (4c)) as a function of the applied current, featuring SNIC spike onset when (a) and homoclinic onset when (b). (c)-(d) Bifurcation diagrams of the model defined in Eqs. (1a) and (1b), with parameter values chosen to mimic the diagrams in (a) or (b), respectively. Limit cycles are shown in purple and fixed points in black, with solid lines for stable branches and dashed lines for unstable branches.

Bifurcation structure and frequency landscape near the SNL bifurcation.

(a) Two-parameter bifurcation diagram versus applied current Iapp of the fast subsystem of the Wang-Buzsáki model (Eqs. (4a) to (4c)) and frequency of the limit cycles. (b) Same as (a), for the QIF model (Eqs. (1a) and (1b)). Note that in this model, the firing frequency goes to infinity as vr approaches vth. Here we only show the frequencies up to a certain distance between those parameters.

When the neuron spikes, to isolate the slow components of the dynamics, in our case overall changes in extracellular potassium concentration, the averaging method can be used to define the averaged slow subsystem (Eq. (6)). The strategy consists in averaging, at given concentrations, the derivative of over one limit action potential cycle of the fast subsystem.

Step 1: QIF version of the Wang-Buzsáki model for a constant

We consider the following QIF model: Its branch of fixed points, defined by Iapp − Isn,0 − Ipump = a(v − vSN)2, folds at (Iapp = Isn,0 + Ipump, v = vSN). Stable limit cycles emerge via a SNIC bifurcation when the reset voltage vr is smaller than vSN (Fig. 3c), while they emerge via a homoclinic bifurcation at when vr > vSN (Fig. 3d), with a SNL bifurcation at vr = vSN. Let us first determine, for constant, parameter values of System (1) such that its bifurcation diagram with respect to the applied current closely resembles the one of the target model at this concentration.

ISN,0 and vSN are simply the coordinates of the saddle-node bifurcation in the target model, in the absence of pump current. Ipump, which in the target model only depends on potassium (Eq. (5d)), can here be treated as a constant negative contribution to the applied current.

We take a and c to be the quadratic normal form coefficients for the saddle-node bifurcation of the target model. They are defined as (see for example [12]): where p and q are the normalized left and right eigenvectors of the Jacobian matrix and B the Hessian quadratic form of the vector field at the saddle-node: Since the bias term c is close to 1 µF cm−2, we set it to this value for simplicity.

Then, if is beyond the value corresponding to the SNL bifurcation in the target model [K+]o,SNL, we choose vr such that the homoclinic bifurcation in the QIF model takes place at the same value Ihom of applied current as in the target model: When we have more flexibility in the choice of reset. vr should be smaller than vsn to obtain a SNIC bifurcation, and a reasonable matching of the limit cycles’ minimum voltage is appreciable. We choose to define the reset in the SNIC case in the continuity of the homoclinic case (see Fig. 4d). This satisfies the two above criteria.

Parameters of the QIF model (Eqs. (1a) and (1b)) depending on the potassium concentration.

(a-d) Derivation of the parameters ISN,0, vSN, a and vr. (e) Bifurcation diagram of the fast subsystem of the target model (Eqs. (4a) to (4c)) with respect to , when Iapp = Iapp,SNL and Ipump = 0 µA cm−2.

Finally, the threshold voltage vth allows us to approximate the limit cycles’ maximum voltage.

Step 2: QIF version of the Wang-Buzsáki model with dynamic

We now compute parameter values of System (1) for a range of , and depending on how strongly they vary with potassium, either set them to a constant or replace them with a function of .

Fig. 3a-d shows ISN,0, vSN and a for between 4 and 16 mM and vr only above [K+]o,SNL. We notice that vSN and the normal form coefficient a are relatively conserved in this interval. We fix them to their value at [K+]o,SNL. For ISN and vr we use linear fits, constrained at [K+]o,SNL, that we extrapolate to lower concentrations in the case of vr. The increase of the reset voltage with reflects the increase of the potassium reversal potential in the conductance-based model.

For the threshold vth, we use a sigmoid, that we extrapolate to lower concentrations, to fit the peak voltage of the limit cycles in the bifurcation diagram with respect to , when Iapp = Iapp,SNL and in the absence of pump (Fig. 3e). This enables us to capture the shrinkage in spike amplitude as potassium accumulates.

We incorporate these functions in System (1) and include potassium dynamics to obtain the following model: with The evolution of the extracellular potassium concentration depends on the one hand on uptake by the Na+/K+-ATPase (Eq. (2b)), and on the other hand on increments at each action potentials (Eq. (2d)). When the target model is Wang-Buzsáki, the parameter values of the corresponding QIF version in the form of (2) are listed in Table 1.

The fast subsystem of this model consists in Eqs. (2a) and (2c), with treated as a parameter. Its bifurcation diagram with respect to input and potassium, and the frequency of the limit cycles, can easily be calculated analytically (see “Methods”). We can see in Fig. 3 that, with our reduction procedure, we achieve a very close match of the bifurcation branches emanating from the SNL bifurcation in the target model. In Wang-Buzsáki, the transition to depolarization block at larger potassium concentrations is mediated by a fold of limit cycles and a subcritical Hopf bifurcation (panel A). The dimensionality of the QIF fast subsystem does not allow for such bifurcations. Instead, we artificially obtain a depolarization block by silencing the dynamics when the threshold voltage vth drops to the reset level (see gray and green curves in Fig. 4e). Note that, although the firing frequency in the fast subsystem grows to infinity as approaches this point, this does not effectively occur in the full system since is incremented by a constant amount with each spike.

The averaged slow subsystem of System (2) is simply given by where T is the period of the limit cycles.

Our phenomenological neuron model closely captures the dynamics of the original conductance-based version

Under identical stimulation, the behavior of the phenomenological model resulting from this bifurcation diagram fitting procedure is very similar to that of the target model. In the same way as the original version, it can produce regular tonic firing (Fig. S2), as well as richer firing patterns. We illustrate this point with three scenarios typically occurring in class I conductance-based model [7, 13, 14], that can notably be obtained in the Wang-Buzsáki model (see overview in Fig. 5). They share the common feature of involving switches in firing regimes, caused by an activity-mediated accumulation of extracellular potassium. Those switches rely on: (i) the different regimes associated with each region of their common bifurcation diagram, (ii) slow (on average) changes in . Although the time traces are not exactly the same, we will see that the two models are quantitatively comparable.

Three scenarios of activity-mediated switches in firing regimes that were reported in class I conductance-based models.

(a-c) Voltage traces, in the case of the Wang-Buzsáki model. (d) Schematic representation of trajectories in the versus applied current bifurcation diagram, through regions associated with various activity regimes.

Scenario 1: fold/homoclinic bursting

We first consider the hysteresis loop bursting mechanism described in [7], which relies on the interplay between slow changes in extracellular potassium concentration and voltage dynamics. It requires only fast-inactivating sodium channels, delayed rectifier potassium channels, and an indispensable electrogenic pump, without the need for any slow ion channels.

Starting from a physiological potassium concentration ( [15]), we stimulate the Wang-Buzsáki neuron with a constant deterministic applied current (Iapp = 0.25 µA cm−2). This corresponds to an initial condition in the spiking region (Fig. 6a), i.e., where the only attractor is a stable limit cycle with large voltage amplitude (Fig. S1a). There, potassium release through voltage-gated channels outweighs the effect of the pump on average over the spike cycles, leading to a progressive accumulation of extracellular potassium (first part of Fig. 6b).

Fold/homoclinic bursting scenario (deterministic input: Iapp = 0.25 µA cm−2).

For the Wang-Buzsáki model: (a) Bursting trajectory shown on the 2-parameter bifurcation diagram [K+]o versus Iapp of the fast subsystem. (b) Voltage and [K+]o time traces. (c) Enlargement of the burst highlighted in panel (b). (d) Bursting trajectory superimposed onto the 1-parameter bifurcation diagram with respect to [K+]o, zooming in on lower voltage values for better visibility. (e) Time derivative of the averaged slow subsystem. (f-j) Same as (a-e), for the integrate-and-fire version.

When the system arrives in the area enlarged in Fig. 6a, where the bistable region is bounded from below by the fold bifurcation (lower ) and from above by the homoclinic bifurcation (higher ), it begins to burst (second part of Fig. 6b). For completeness, note that this configuration is only a necessary condition for the bursting behavior to occur (for more details, see [7]). During each spiking phase, the system follows the branch of stable limit cycles (Fig. 6d), in the course of which potassium increases overall (Fig. 6e) because the increments at each action potential are not fully compensated by the pump (Fig. 6c). When it reaches the homoclinic bifurcation, the system switches to the branch of stable equilibria, initiating a quiescent phase. In the absence of spiking, the pump can restore potassium concentration down to the fold bifurcation, at which point the system jumps back to the stable limit cycle branch. This marks the start of a new bursting cycle.

Starting from the same initial voltage and potassium concentration, and stimulated with the same input, the integrate-and-fire version of the model successfully captures this bursting behavior (Fig. 6f-j). The underlying mechanism is the same as in the original model, with fold and homoclinic dynamical bifurcations governing the onset and termination of bursts. The small discrepancy between the two models in the number of spike per burst and interspike interval is due to the divergence in the frequency of limit cycles (Fig. S1d) and slightly different location of the bifurcations in relation to potassium levels.

We should point out here that the intersection between the x-axis and the time derivative of the averaged slow subsystem, visible in Fig. 6e,j, does not correspond to a stable limit cycle of the full system. Although there exists a potassium concentration for which, in the fast subsystem, the contributions of the pump and voltage-gated channels to potassium dynamics cancel each other out on average, in the full system the hysteresis loop forms and bursting takes place. As explained in [16], this results from the fact that the reduction of the dynamics to the averaged slow subsystem is not valid in the close vicinity of the homoclinic bifurcation.

For this first scenario, in accordance with [7], we adopted a configuration where the potassium concentration rises until the neuron starts bursting. However, note that simulations can also be generated in which the neuron remains in the spiking region, for example by implementing a buffering of the extracellular potassium. This can conveniently be done in both original and reduced versions of the model, by introducing a term in the potassium equation. As a result, when the buffering is sufficiently strong, potassium stabilizes within the spiking region at a concentration corresponding to a fixed point of the average slow subsystem which, in this case, equates to a limit cycle of the full system (Fig. S2).

Scenario 2: stochastic bursting

We now examine a switch in firing regime identified in a version of the Traub-Miles model [17] (a class I conductance-based model) accounting for potassium concentration dynamics. By introducing a noise component in the input, Contreras et al. obtained simulations showing that, after a period of tonic firing, the neuron switches to stochastic bursting, a regime where firing is intermittently interrupted [13]. It is possible to reproduce this behavior with the Wang-Buzsáki model.

We stimulate the neuron with colored noise (see Eq. (8) in “Methods”) of mean input Iapp = 0.3 µA cm−2, chosen sufficiently large to not interfere with the fold/homoclinic bursting behavior (Fig. 7a). Initially, like in the previous scenario, the neuron spikes periodically, causing an accumulation of extracellular potassium (first part of Fig. 7b). When the system reaches the bistable region, i.e., when crosses the fold bifurcation of the fast subsystem, the firing pattern switches to stochastic bursting (second part of Fig. 7b). There, the noisy input triggers jumps between the two attractors: the stable fixed point corresponding to a resting state, and the stable limit cycle corresponding to tonic firing (Fig. 7d).

Stochastic bursting scenario (noisy input, with mean Iapp = 0.3 µA cm−2).

For the Wang-Buzsáki model: (a) Trajectory shown on the 2-parameter bifurcation diagram [K+]o versus Iapp of the fast subsystem. (b) Voltage, [K+]o and applied current time traces. (c) Enlargement of the burst highlighted in panel (b). (d) Part of the trajectory corresponding to the burst highlighted in panels (b) and (f), superimposed onto the 1-parameter bifurcation diagram with respect to [K+]o. (e-h) same as (a-d), for the integrate-and-fire version.

As expected from their common bifurcation structure, when confronted with the same stimulus (same realization of the stochastic process) as the target model, the integrate-and-fire version behaves very similarly (Fig. 7e-h).

Compared to the deterministic fold/homoclinic bursting, for which burst initiation and termination occur only at the boundaries of the bistable region, in this scenario the noise can induce switches between attractors within the bistable region (Fig. 7c-d,g-h).

On a side note, we notice a canard-like behavior in the middle of the burst shown in Fig. 7c-d, where for a short duration the system remains near the saddle point (unstable).

Scenario 3: transition to depolarization block

The last scenario we consider is the transition to depolarization block, which results in transient spike inactivation. This phenomenon is a pathological hallmark: it is the signature of cortical spreading depolarization at the single-neuron level, and might also be involved in epileptiform activity [18]. The induction of a depolarization block, in the case where it is caused by an accumulation of extracellular potassium resulting from neuronal firing, has been extensively studied in conductance-based models with dynamic ion concentrations (e.g. [13, 14, 19–22]). We simulate it here with the Wang-Buzsáki model.

To achieve this, we use the same setup as for the stochastic bursting scenario, but without the noise component in the stimulus. As a result, in the bistable region the system remains on the branch of stable limit cycles, allowing the extracellular potassium concentration to build up without interruption. When exceeds the concentration corresponding to a fold of limit cycles of the fast subsystem, the neuron enters depolarization block (Fig. 8a-d). In comparison, in the stochastic bursting scenario, frequent pauses in firing due to the presence of noise make it easier for the regulatory mechanisms—in our model the Na+/K+-ATPase—to maintain in the bistable region, preventing the pathological block.

Depolarization block scenario (deterministic input: Iapp = 0.3 µA cm−2).

For the Wang-Buzsáki model: (a) Trajectory shown on the 2-parameter bifurcation diagram [K+]o versus Iapp of the fast subsystem. (b) Voltage and [K+]o time traces. (c) Enlargement of the transition to depolarization block (b). (d) Part of the trajectory highlighted in panels (b) and (f), superimposed onto the 1-parameter bifurcation diagram with respect to [K+]o. (e-h) Same as (a-d), for the integrate-and-fire version.

When stimulated with the same deterministic input, the integrate-and-fire version of the model produces a voltage trace very similar to the depolarization block observed in the Wang-Buzsáki model (Fig. 8f-g). However, unlike the two other scenarios, this behavior does not entirely rely on the same mechanism as in the original model. The path leading to the block is unchanged, but the transition to the block itself cannot occur via the same dynamical bifurcation, due to the design of the reduced model. Instead, we emulate it by placing the system (both the voltage and ) in refractory period as soon as, at larger values, the reset voltage meets the threshold value (Fig. 8e,h). With this constraint, although the dynamics during the block is only sketchily captured, the desired outcome is for the neuron to ceases spiking. The dynamics preceding the block is, for its part, well captured: the QIF voltage trace features both the reduction in spike amplitude and increase in firing frequency (Fig. 8g). These two effects are in agreement with the experimental recordings in [20]. As shown in Fig. S1f,h, the frequency of the limit cycles of the fast subsystem as a function of increases in two stages before the block. This pattern is a remnant of the situation with two homoclinic bifurcations, at lower values of applied current (Fig. S1b,d).

Reshaping of the phase-response curve

We now compare the phase-response curves (PRCs) of the two models. For a spiking neuron, the PRC describes how the phase of the cycle at which a perturbation is received affects the timing of the next action potential. It provides a link between the dynamics of single neurons and behavior at the network scale.

Fig. 9a shows the infinitesimal PRC (iPRC, i.e., for infinitesimal input amplitude) of the Wang-Buzsáki model near spike onset, for different potassium concentrations. Like when the SNL is reached via other biophysical parameters (e.g. see [24] for the capacitance), we can see a transition from almost symmetric at low [K]+ to left-tilted PRCs after the SNL bifurcation. Near spike onset, the alteration of the iPRC occurs suddenly at the SNL, as illustrated with the cases and note that for the latter two, the curves are nearly identical.

Effect of on the phase response curve and ability to synchronize, near spike onset .

(a) iPRCs for the Wang-Buzsáki model (direct method). (b) Spike voltage traces for the Wang-Buzsáki model. (c) iPRCs for the integrate-and-fire model (analytically, see “Methods”). (d) Spike voltage traces for the integrate-and-fire model. (e) Period of the limit cycles. (f) Locking range, assuming delta coupling with a voltage perturbation of 0.15 µV. For the computation of the coupling function, pulses were normalised following Weerdmeester et al. [23].

The PRC reshaping at the SNL is well captured by the QIF model (Fig. 9c). In fact, this simplified model offers an intuitive understanding of how an increase in deforms the iPRC. The effects of larger are mainly expressed in the QIF model by a larger reset voltage, which results in skipping the beginning of the spike cycle (Fig. 9d). Since we are dealing here with a one-dimensional system, the iPRC is simply the inverse of the derivative of the spike voltage trace as a function of the phase. Increasing the reset is therefore equivalent to retaining only a scaled version of the iPRC segment corresponding to late phases (Fig. S3). As pointed out in [24], the SNL bifurcation coincides with a reduction of the period at a given distance from spike onset (i.e., sharper f-I curves), resulting in larger iPRC amplitudes (Fig. 9a,c). This period reduction is captured by the QIF version of the model (Fig. 9e).

The changes associated with the SNL bifurcation directly affect the collective behavior of neurons in a network. Based on the theory of weakly coupled oscillators [25], and given a specific network configuration, the iPRC can be used to predict the neurons’ ability to synchronize. For example, in the case of two neurons coupled via weak delta synapses, the coupling function is a scaled version of the iPRC, and its odd part governs how the phase difference between the two neurons evolves over time [23, 26, 27]. When the uncoupled firing frequency differs between the two neurons, the locking range, i.e., the maximal frequency detuning such that there exists a stable fixed point for the dynamics of the phase difference, is twice the amplitude of the odd part of the coupling function [26]. Fig. 9f shows the locking range as a function of potassium, for a voltage perturbation amplitude of 0.15 µV. For both models, the locking range abruptly increases at the SNL, indicating enhanced synchronization ability. In more complex networks, the analysis must be adapted to the particular network architecture (e.g. network topology, synaptic coupling), but the PRC can still serve to predict aspects of the network state [28].

Extracellular potassium influences neurons synchronization in a network

We now investigate the consequences of the changes introduced by the SNL bifurcation at the network scale. In particular, for tonically spiking neurons, we expect that the left-tilted PRCs at larger potassium concentrations induce in-phase synchronization of inhibitory networks [29]. We test this prediction in a basic network configuration. For this, we consider 100 inhibitory neurons, fully coupled with delta synapses. We then impose on all these neurons a common extracellular potassium trace, simulating for instance an external wave reaching the network (Fig. 10a).

Response of a network of 100 neurons to an external potassium wave.

We considered an all-to-all connectivity scenario, with weak inhibitory pulse-coupling (δv = −0.000 15 mV). We used a noisy input (see “Methods”), common to all neurons, with mean Iapp = 0.25 µA cm−2. (a) Incoming potassium wave, common to all neurons. (b) 2-parameter bifurcation diagram versus Iapp of the fast subsystem, for the Wang-Buzsáki model. The crosses indicate the initial and final potassium concentrations. (c-d) Raster plot for 10 among the 100 neurons, during the third and 13th seconds, respectively. (e-g) Same as (b-d), for the QIF version of the model.

With both models, the neurons of the network, which were previously out of synchronization (Fig. 10c,f), become in-phase synchronized upon arrival of this wave (Fig. 10d,g).

Disentangling the respective contributions of PRC asymmetry and frequency fluctuations to the effect of the SNL transition on synchronization requires further investigation and lies beyond the scope of this study. Furthermore, we deliberately chose values such that the neurons remain in the tonic spiking region (Fig. 10b,e). Indeed, in the bistable region, where only a fraction of the neurons may be active, the PRC is not a reliable predictor of network behavior. A study of the complex dynamics at work in this region is beyond the scope of this article (see for example [30]).

Discussion

Neuronal dynamics can influence network behaviour, yet the inclusion of biophysical detail in larger network models, both challenges computational efficiency and reduces model generality due to the amount of additional parameters introduced. Commonly, neuron models are fitted to datasets from a single experimental setting and one best-fit parameter set is determined. These models may fail to generalize, if parameters not varied in the experiment are altered in the model. We here derive models that capture the impact of a biophysical parameter across a larger parameter space by fitting not voltage data but bifurcation diagrams. The resulting model captures the dynamical transitions in parameter space and therefore, generalises at the cost of over-fitting at a particular parameter value.

At the scale of a single neuron, we have shown that the phenomenological model accurately reproduces various activity patterns from the original model, without needing to adjust parameter values to each scenario. At the network level, preliminary simulations highlight the important role that extracellular potassium concentration plays in synchronization.

Specifically, we have here shown how to reduce a class I conductance-based model with dynamic potassium concentration to a quadratic integrate-and-fire model. Depending on the concentration of extracellular potassium and its dynamics, single-cell excitability in this model can vary between the SNIC or the HOM type, capturing the transitions in the full conductance-based neuron models. The effect of potassium on the bifurcation structure in the original, conductance-based model is in the reduced QIF model version implemented via a change in the voltage reset variable after a spike. Note that this approach is not limited to potassium; it could be extended to other parameters known to induce an SNL transition (e.g. temperature [26] or capacitance [24]).

A phenomenological neuron model capturing potassium dynamics

Phenomenological models that capture biophysical aspects are particularly valuable for exploring aspects at the network scale through systematic simulations. For example, phenomenological models of spike-frequency adaptation have been used to study network synchronization [2, 31]. In the same vein, our work formulates a simple model of potassium concentration dependence. It provides a minimal description of a dynamics that would typically require a more detailed, higher-dimensional system.

Extracellular potassium concentration is a particularly pertinent variable to retain in simplified neuron models. Originally, models assumed constant ion concentrations. This is motivated by the fact that the amount of relocated ions per action potential may have a negligible effect on the overall concentrations [32, 33], and that concentrations are regulated by mechanisms involving the Na+/K+-ATPases, cotransporters and glial cells [34, 35]. Nonetheless, in the last decades numerous studies have focused on modeling their dynamics [7, 8, 13, 14, 22, 36–43]. Indeed, experimental studies suggest activity-dependent fluctuations in concentration, in particular transient accumulations of extracellular potassium reaching several millimolars, in case of intense activity or following sensory stimulation [6, 34, 35, 41, 44, 45]. Such changes are expected to be particularly significant locally or in restricted spaces [34, 35]. Moreover, drastic increases in extracellular potassium are known to occur during pathophysiological events such as spreading depolarizations and epileptic seizures, as well as during hypoxia [34]. Both experimental evidence [13, 45–47] and results based on biophysical neuron models (e.g. [7, 8, 13, 14, 22, 36, 37, 43]) indicate that the weakening of the potassium driving force, resulting from activity-mediated extracellular build-up, in turn qualitatively alters neuronal excitability, at both moderate and large elevations. This motivates the need for simple models capturing potassium dynamics, to investigate the consequences of changes in neurons’ intrinsic excitability at the network scale.

There exist other minimalist models with potassium dynamics, mainly in the context of epilepsy [48–50]. The single neuron model by Depannemaecker et al. [48] reproduces a wide range of firing patterns, including seizure-like events, sustained ictal activity and depolarization block, while being relatively simple. It remains nevertheless at a higher level of complexity compared to the present model, with its two-dimensional fast subsystem of type Hodgkin-Huxley and its two slow variables. The “epileptor-2” population model [49] by Chizhov et al. is also four-dimensional, with the advantage of explicitly modeling sodium dynamics. It can successfully simulate interictal and ictal discharges in high potassium conditions. However, the choice of coupling between potassium and voltage dynamics in this model does not permit onset bifurcation switches in individual neurons. In a recent study, Depannemaecker et al. propose an extension of the adaptive exponential integrate-and-fire model (AdEx) that accounts for impaired ionic gradients [50], by introducing a third, slow, variable.

Modeling the critical transitions in parameter space

The question of “what is a good model” has been debated extensively in the epistemological literature and many arguments in favor of simple leaky integrate-and-fire models have been presented [51], for example, if well fitted, they have high predictive power for the voltage state variable. However, if one includes model predictions not only in observable state variables, but also for changed biophysical parameter sets, then only models that share the fundamental bifurcation structure found in the data yield good predictions. Our work, therefore, presents a strategy to derive models that capture the dependency of the system on physiological quantities, by matching the relevant bifurcation diagram. For a related approach in a broader context, see [52]. This way of constructing phenomenological models is a real alternative to approaches based on the fitting of spike trains. It offers advantages that can be particularly beneficial, depending on what the model is meant to achieve.

This is of additional relevance since nerons are not stationary devices, but subject to constant change in their enviromnetal parameters. Sources of variability include temperature [26, 53], ionic concentrations [13, 14] and neuromodulators [54]. These factors may directly depend on the recent history of activation [13, 55]. In some cases, their changes constitute a homeostatic response to other changes, helping to maintain the neurons’ properties in the face of a varying environment [56]. We argue that, when known, fitting the bifurcation diagram with respect to chosen parameters is a judicious way to obtain models tailored to the major source of instability in a given context, as we have done here with the extracellular potassium concentration. Fitting specific spike trains typically ensures model validity only in the idiosyncratic conditions of their construction. In contrast, fitting a bifurcation diagram not only provides better robustness to changing conditions, but also allows for the study of how these changes themselves slowly drive neuronal activity between different regimes.

There already exist canonical phenomenological models that, thanks to their bifurcation landscape, can recapitulate the main features of neural excitability (see for example [57], or [58], which is based on the unfolding of a higher-dimensional bifurcation). However, they are formulated in terms of abstract variables. In some situations, having a model that contains specific bifurcations is not sufficient; their arrangement in terms of biophysical dimensions also matters. For example, the fold/hom bursting mechanism that we studied in the first scenario cannot occur when the unfolding of the SNL bifurcation is aligned with the axis (see [7] for more details). Starting from a canonical model and adapting it to match desired bifurcation branches, as we did here with the QIF model, has the advantage of preserving important quantitative components.

Potassium dynamics and network models

One of the possible applications of our work is to provide a tool for studying the consequences of potassium-induced excitability switches at the network level. By adding potassium diffusion to the network model future studies could explore spatial effects and the interplay between synaptic and diffusive coupling (see [59, 60]).

Possible model extensions

Sodium dynamics

In this work, we restricted ourselves to potassium dynamics, although sodium concentrations are also directly affected by neuronal activity. This choice was made because potassium is the main player in activity-induced SNL transitions and switches in excitability class, while sodium is more involved in spike amplitude reduction [13]. The engagement of the Na+/K+-ATPase to restore ionic gradients is accounted for in our model by the potassium dependence of the pump current. In [7], it is shown that sodium dynamics is not crucial for the bursting mechanism we studied in the first scenario.

Taking into account the dynamics of the two ion species, for example by tracking extracellular potassium and intracellular sodium concentrations, remains the most precise option [49, 61]. To avoid adding an extra variable, one could rely on electroneutrality to express one concentration as a function of the other [21, 36, 48]. Applying this strategy to our work is an interesting direction for refinement.

Transition to depolarization block

Finally, the way in which depolarization block is obtained in the reduced model (Scenario 3) is not ideal. Instead of being mediated by a supercritical Hopf bifurcation, or a subcritical Hopf coupled to a fold of limit cycles, as in conductance-based models, it was artificially achieved by silencing the dynamics when the reset and threshold voltages meet. Indeed, the fast subsystem of the reduced model, of type quadratic integrate-and-fire, cannot exhibit Hopf bifurcations. Instead, one would need to start from a two-dimensional fast subsystem, like the adEx model. This would not only allow a more accurate modeling of the transition to the block, but also open the possibility of studying the recovery from this pathological state.

Conclusion

In conclusion, we affirm the importance of phenomenological models that capture key underlying biophysical mechanisms. We contribute a model of potassium concentration dependence, readily suitable for integration into large networks. Moreover, our approach, based on the fitting of a bifurcation structure, provides a generalizable framework for deriving similar models for other relevant modulatory variables.

Methods

Extended Wang-Buzsáki [7] model

Model definition

The model obeys the system of differential equations with currents and gating functions The reversal potential for the potassium evolves with according to where R = 8314.4 mJ K−1 mol−1 is the ideal gas constant, F = 96 484.6 C mol−1 the Faraday’s constant and T = 310 K the temperature. The variables and parameter values are listed in Tables II and III.

Slow dynamics

As in [7], for values such that there is a stable limit cycle in the fast subsystem, we define the averaged slow subsystem of System (4) as where T is the period of limit cycles.

Fast subsystem of the QIF model (Eqs. (2a) and (2c))

Frequency of the limit cycles

Let (tonic spiking region), the period of the limit cycles is given by: If vr > vth and (bistable region): with

Infinitesimal phase-response curve (iPRC)

Let . If : If vr > vth and :

Noisy input

To introduce noise in the applied current we replace it with: where ηt is a Ornstein-Uhlenbeck (OU) process defined by Wt being the Wiener process. The OU process can be interpreted as a smoothly low pass filtered white noise. In our simulations, we used a correlation time τη = 20 ms and a standart deviation σ = 0.05 µA cm−2.

Computational resources

To derive the parameters ISN,0, vSN, a and vvr of the QIF model (1), we performed symbolic computations with the Python library Sympy [63]. In the case of vsn, we additionally used the numerical solver nsolve.

For numerical continuation, we used the software AUTO-07P [64].

Simulations of the neuron models were carried out with the python package brian2 [65].

eLife Assessment

This work demonstrates an objective way to select parameter values for a quadratic integrate-and-fire model so that its bifurcation diagram matches a specific target diagram, generated from the Wang-Buzsaki model. The method is useful for the field and is presented with convincing evidence. The method is currently limited in its ability to be applied to data, but improves our mathematical tools to treat a rarely studied type of bifurcation.

Reviewer #1 (Public review):

Summary:

From a big picture viewpoint, this work aims to provide a method to fit parameters of reduced models for neural dynamics so that the resulting tuned model has a bifurcation diagram that matches that of a more complex, computationally expensive model. The matching of bifurcation diagrams ensures that the model dynamics agree on a region of parameter space, rather than just at specially tuned values, and that the models share properties such as qualitative features of their phase response curves, as the authors demonstrate. A notable point is the inclusion of extracellular potassium concentration dynamics into the reduced model - here, the quadratic integrate-and-fire model; this is straightforward but nonetheless useful for studying certain phenomena.

Strengths:

The paper demonstrates the method specifically on the fitting of the quadratic integrate-and-fire model, with potassium concentration dynamics included, to the Wang-Buzsaki model extended to include the potassium component. The method works very well overall in this instance. The resulting model is thoroughly compared with the original, in terms of bifurcation diagrams, production of various activity patterns, phase response curves, and associated phase-locking and synchronization properties.

Weaknesses:

It is important to note that the proposed method requires that a target bifurcation diagram be known. In practical terms, this means that the method may be well suited to fitting a reduced model to another, more complicated model, but is not likely to be useful for fitting the model to data. Certainly, the authors did not illustrate any such application. Secondly, the authors do not provide any sort of general algorithm but rather give a demonstration of a single example of fitting one specific reduced model to one specific conductance-based model. Finally, the main idea of the paper seems to me to be a natural descendant of the chain of reasoning, starting from Rinzel - continuing through Bertram; Golubitsky/Kaper/Josic; Izhikevich; and others - that a fundamental way to think about neuronal models, especially those involving bursting dynamics, is in terms of their bifurcation structure. According to this line of reasoning, two models are "the same" if they have the same bifurcation structure. Thus, it becomes natural to fit a reduced model to a more complicated model based on the bifurcation structure. The authors deserve credit for recognizing and implementing this step, and their work may be a useful example to the community. But the manuscript should have described and cited this chain of works to put the current study in the correct context.

Reviewer #2 (Public review):

Summary:

The authors derive an integrate-and-fire model to describe the dynamics of a more complex Wang-Buzsaki model and compare the two models. A detailed discussion of bifurcation schemes in both models is convincing and allows us to evaluate the simpler model.

Strengths:

The idea is interesting, and the mathematical approach appears to be convincing. In addition, differences between the simple and original models are also discussed.

Weaknesses:

A comparison to experimental data is necessary to support the theoretical work.

Author response:

Public Reviews:

Reviewer #1 (Public review):

Summary:

From a big picture viewpoint, this work aims to provide a method to fit parameters of reduced models for neural dynamics so that the resulting tuned model has a bifurcation diagram that matches that of a more complex, computationally expensive model. The matching of bifurcation diagrams ensures that the model dynamics agree on a region of parameter space, rather than just at specially tuned values, and that the models share properties such as qualitative features of their phase response curves, as the authors demonstrate. A notable point is the inclusion of extracellular potassium concentration dynamics into the reduced model - here, the quadratic integrate-and-fire model; this is straightforward but nonetheless useful for studying certain phenomena.

Strengths:

The paper demonstrates the method specifically on the fitting of the quadratic integrateand-fire model, with potassium concentration dynamics included, to the Wang-Buzsaki model extended to include the potassium component. The method works very well overall in this instance. The resulting model is thoroughly compared with the original, in terms of bifurcation diagrams, production of various activity patterns, phase response curves, and associated phase-locking and synchronization properties.

Weaknesses:

It is important to note that the proposed method requires that a target bifurcation diagram be known. In practical terms, this means that the method may be well suited to fitting a reduced model to another, more complicated model, but is not likely to be useful for fitting the model to data. Certainly, the authors did not illustrate any such application. Secondly, the authors do not provide any sort of general algorithm but rather give a demonstration of a single example of fitting one specific reduced model to one specific conductance-based model.

We thank the reviewer for this critical assessment. It is true that we demonstrate our approach using as target the bifurcation diagram of a more realistic model. In principle, the method would be applicable to experimental systems if parameter space is sampled under appropriate experimental conditions, e.g., by recording at different extracellular potassium concentrations. However, this is challenging: generating reliable” experimental bifurcation diagrams” would require tightly controlled experimental repetitions under various conditions, particularly for reconstructing two-dimensional bifurcation diagrams. We hope that this work encourages the development of such methods. We have added a discussion of this point; see the paragraph starting at line 549.

We have also included a general algorithm (Table I) and provide a second example illustrating the procedure (Supplementary Fig. S2).

Finally, the main idea of the paper seems to me to be a natural descendant of the chain of reasoning, starting from Rinzel - continuing through Bertram; Golubitsky/Kaper/Josic; Izhikevich; and others - that a fundamental way to think about neuronal models, especially those involving bursting dynamics, is in terms of their bifurcation structure. According to this line of reasoning, two models are “the same” if they have the same bifurcation structure. Thus, it becomes natural to fit a reduced model to a more complicated model based on the bifurcation structure. The authors deserve credit for recognizing and implementing this step, and their work may be a useful example to the community. But the manuscript should have described and cited this chain of works to put the current study in the correct context.

We have added a paragraph in the Discussion section (starting at line 517) to better situate the manuscript within the relevant literature and to explicitly acknowledge the chain of work.

Reviewer #1 (Recommendations for the authors):

Please see my public review. In line with my comments, I recommend that the authors either (a) provide a general algorithm for fitting at least a class of reduced models (i.e., those that satisfy some general assumptions) to a class of bifurcation diagrams, or (b) provide at least one more example of implementing their method. Step (b) would not need to be done to the same degree of thoroughness as the example they provided (e.g., the PRCs and synchrony need not be considered), but to me, this step would be very important if (a) is impractical. Otherwise, the paper should probably be rewritten to de-emphasize the message that this is a general method; instead, this should be a paper about specifically fitting the QIF (with potassium dynamics) to the Wang-Buzsaki model (with potassium dynamics).

We provide a general algorithm for deriving a quadratic integrate-and-fire model with dependence on a biophysical parameter by fitting the bifurcation structure of a given class I conductance based neuron model near an SNL bifurcation induced by this parameter; see Table I.

In addition, we provide a second example of the reduction procedure: motivated by Hesse et al. (Nature Communications, 10.1038/s41467-022-31195-6, 2022), we derive a QIF model that captures dependence on temperature instead of potassium concentration; see Supplementary Fig. S2.

Not surprisingly, I also think it’s essential that the authors describe and cite the chain of works on thinking of neuronal models in equivalence classes based on bifurcation diagrams, and make clear that this paper builds on the ideas set forth in that chain.

We thank the reviewer for this comment. As mentioned above, we have added a paragraph in the Discussion section, starting at line 517, to acknowledge this chain of work.

Also, the authors should make clear that their method is not one for fitting a model directly to data, which will require rewriting at least the first paragraph of their Discussion section.

We thank the reviewer for helping us make our manuscript clearer. To avoid confusion, we have clarified this point already in the Introduction (see lines 52-57) and have included a new paragraph in the Discussion (starting at line 549).

Other specific corrections are:

(1) Typos should be fixed, as the paper has several. The first line of the abstract has one (“concentrations” → “concentration”), for starters. “Original model” on pg. 3 is missing “be” in “can defined”. “ceases” → “cease” on pg. 13. “nerons” → “neurons” on pg. 19. “standart” → “standard” pg. 24.

Done. Additional typos were also corrected.

(2) The abstract mentions “consequences in networks” in its second sentence. This is misleading because studying network dynamics is not at all the main emphasis of the paper, but rather a corollary application of the main ideas, so some restructuring of the abstract is needed. Similarly, the final abstract sentence overstates somewhat what was done with studying synchronization and should be rewritten more precisely.

We have restructured the abstract accordingly.

(3) For readers who are interested in the ideas here but not familiar with the QIF model, it will be very difficult to follow the first paragraph of Results. Elementary aspects of QIF dynamics should be explained here (e.g., what is the saddle-node bifurcation), and a basic figure panel about this should be included in Figure 1.

We added a supplementary figure (Figure S1) adapted from Izhikevich for readers who might not be familiar with the QIF model.

(4) Bottom lines of page 3 should be reworded to make clear that the slow variables are averaged over each member of a family of fast subsystem limit cycles. Also, “one limit action potential cycle” is an awkward phrase.

We rephrased this sentence (see paragraph starting at line 140).

(5) Text under system (1) – why isn’t c mentioned? Also, references to Figure 3 should be to Figure 2 here. And authors should state what they mean by “target model” and be clear about whether it includes potassium dynamics and/or pump current.

c scales the parabola corresponding to the branch of fixed points, given by c(Iapp − ISN,0 − Ipump) = −a(ν − νSN)2. Consequently, it also affects the position of the homoclinic bifurcation: Iapp=ISN,0+Ipump−ac(vr−vSN)2 In the previous version of the manuscript, c was inadvertently omitted from the expression for the branch of fixed points; this has now been corrected. See paragraph starting at line 154.

Figure references have been corrected.

By target model, we mean the conductance-based model that includes potassium dynamics and a pump current, in our case System 4. We clarified this point at the beginning of the Results section (see paragraph starting at line 116). Throughout the manuscript, we now explicitly indicate when we refer only to its fast subsystem and whether the pump current is included. In particular, note that the parameter derivation shown in Figure 4 is performed on the fast subsystem of the target model, in the absence of the pump current. Ipump can be considered as a potassium-dependent contribution to the applied current, and can be added a posteriori to the QIF model. See paragraph starting at line 162.

(6) Next par: is the “saddle-node bifurcation” that with Iapp as bifurcation parameter? Please clarify.

Yes, it is the saddle-node bifurcation with Iapp as bifurcation parameter. We have clarified this in the manuscript; see the paragraph starting at line 170.

(7) Bottom pg. 5: does “beyond” mean above? below?

We meant above (larger values of [K]o+). In the text, we have replaced “beyond” with “larger than”. See paragraph starting at line 178.

(8) Formula for vr at top of page 6: Please specify what formula for Ipump is being used here.

The formula for Ipump is given in Eq. 5d. We are using the same formula throughout the paper.

Note that to clarify the reduction procedure, we derive QIF parameters to match the bifurcation diagram with respect to the applied current of the fast subsystem of the target model when Ipump = 0. Reintroducing Ipump produces the same horizontal shift in this bifurcation diagram for both the QIF and Wang–Buzsáki versions.

We have restructured the paragraph starting at line 178 to clarify these aspects.

(9) Three lines below this: I don’t understand what “matching...is appreciable” and “in the continuity of...” mean. Please revise and also explain why a closer matching of vr to the min voltages in Figure 4d was not used, and exactly how the vr that is shown was chosen.

With “matching...is appreciable”, we meant that values assigned to vr should be close to the minimum voltage values reached during spiking. With ”in the continuity of...”, we meant that when [K]o+<[K+]o,SNL.(SNIC case), we choose vr by extrapolating the linear fit performed on the values of vr assigned when [K]o+<[K+]o,SNL, (homoclinic case).

When [K]o+>[K+]o,SNL, vr was chosen so that the homoclinic bifurcation occurs at the same value of applied current as in the fast subsystem of the target model. This is explained in the paragraph starting at line 178 (see Eq. 2). This criterion also allows the minimum voltage values reached during spiking to be captured reasonably well (compare the green dotted line and the purple solid line in panel e of Figure 4).

We have rewritten the paragraph starting at line 186 to clarify these aspects.

(10) Bottom page 6 - reference to Figure 3e should be 4e. Also, the text mentions the shrinkage of spike amplitude, but the figure shows that vth increases over most of the K+ range before decreasing, so a correction is needed.

We corrected the figure reference.

The maximal voltage of the limit cycles of the target model’s fast subsystem (upper purple curve in Figure 4e) increases slightly between [K]o+=[K+]o,SNL≈9.3mM and [K]o+=[K+]o,SNL≈9.3mM, by less than 1mV. It then decreases by about 27mV before the fold of limit cycles. The sigmoidal function vth ([K]o+=[K+]o,SNL≈9.3mM) allows us to capture this substantial decrease in the QIF model. The small preceding increase is not captured. We reformulated the text to avoid confusion (see paragraph starting at line 209).

(11) Figure 4d: Why is EK plotted here? It should be mentioned in the caption and text. More generally, the caption for Figure 4e should be expanded to mention what the purple curves are, what is the black curve for K < KSNL, and what the other structures shown are. Finally, the text describes that theSNIC/SNL/Hom is determined by the choice of vr relative to vSN, so it’s not clear what is Iapp,SNL in the caption - please clarify.

In conductance-based models, higher [K]o+ weakens the potassium concentration gradient, thereby raising EK. The sodium and potassium reversal potentials typically bound voltage oscillations during spiking (see for example Chander and Chakravarthy, PLOS ONE, 10.1371/journal.pone.0048802, 2012), so the minimum voltage of spikes is expected to be higher when [K]o+ is larger. We plotted EK in Figure 4d to show that the increase of the reset voltage vr at larger [K]o+ reflects this effect in the QIF version of the model. We clarified this in the caption of Figure 4 and in the text (see paragraph starting at line 199).

Purple curves show families of limit cycles, while black curves, including the one for [K]o+<[K+]o,SNL, show families of fixed points. The green curve shows the linear fit of vr from panel d. All these have now been included in the legend.

In the QIF model, the onset bifurcation (SNIC, SNL, or homoclinic) is indeed determined by the choice of vr relative to vSN. Figure 4e shows the bifurcation diagram of the fast subsystem of the conductance-based model (Wang-Buzsáki). This is a bifurcation diagram with respect to [K]o+, for a fixed value of applied current. We chose to fix Iapp at its value at the SNL bifurcation, denoted Iapp,SNL. Iapp,SNL is near 0.22 (see Figure 3a).

(12) Eqn. (2a): Shouldn’t vSN depend on potassium like ISN and Ipump do? What is the formula for Ipump there? Why isn’t the RHS of (2b) dependent on v as in the original model? Please clarify.

For simplicity, we did not include a potassium dependence for vSN in the QIF model. Instead, we set it to its value at the SNL bifurcation (Figure 4b). This is explained in the paragraph starting at line 199: “We notice that vSN and the normal form coefficient a are relatively conserved in this interval. We fix them to their value at [K+]o,SNL.”

The formula for Ipump is given in Eq. 5d. The potassium dynamics depends on the voltage via the reset rule in Eq. 3d. This allows us to capture the small increments in [K]o+ at each action potential in the original model (see for example Figure 6c,h). We have clarified these two points in the manuscript (see paragraph starting at line 223).

(13) Figure 3: Please indicate the criticality of the Hopf bifurcations shown.

The legend of Fig. 3 now indicates that the Hopf bifurcations are subcritical. The same clarification has been added to the following figures as well.

(14) Bottom pg. 9: Why isn’t there an Ipump term as in eqn. (6a)? Please clarify.

You are correct, the Ipump term should be included in the equation for the averaged slow subsystem of the QIF model (see paragraph starting at line 243); it was accidentally omitted in the manuscript. Thank you for pointing this out.

(15) Top pg. 11: It’s important to reference the slow averaged dynamics here, which allows K+ to increase. Also, this first paragraph should already explain that this averaged dynamics is only relevant along the family of FS periodic orbits, not during the recovery when the FS has a branch of stable equilibria.

The averaged slow subsystem is indeed only relevant along families of limit cycles of the fast subsystem. Along families of equilibria, averaging is not necessary and the standard slow subsystem can be used. We now explicitly define this standard slow subsystem for both Wang-Buzsáki and the QIF model (see paragraphs starting at lines 241 and 628). In panels e and j of Figure 6, both systems are now represented.

In the paragraph starting at line 273, we now refer to the averaged slow subsystem to explain the overall increase of [K]o+ during bursts (purple curves in Fig. 6e,j), and to the standard slow subsystem to explain the decrease of [K]o+ during quiescent phases (black curves).

(16) Pg. 11, par 3: This is unnecessarily confusing. Please try to reword and clarify this paragraph.

We have simplified this paragraph (starting at line 293). The key point is that the reduction to the averaged slow subsystem is not valid too close to the homoclinic bifurcation.

(17) Pg. 13, end of Scenario 2: Is there any evidence this is a canard effect and not a noise effect? If so, please mention the evidence; otherwise, perhaps take this out.

What happens there appears to be a noise-induced canard effect: in the beginning of the burst, the system follows a family of stable limit cycles of the fast subsystem, i.e. a stable object. However, at some point, noise induces a transition to a portion of trajectory where the system evolves near the saddle branch, i.e. a repelling object, for a substantial amount of time. Such phenomena have been thoroughly investigated in the literature, and can also be obtained in a deterministic way; see for example Marin et al. (Physical Review E 90, 042718, 2014). Bursting traces similar to the one in Figure 7c are observed experimentally (see, for example, Figure 4c of Marin et al.), which is why we considered it worth mentioning. We have revised the paragraph starting at line 332 to clarify this point.

(18) I only see 4 curves in Figure 9a,c, but the legend has 5. Are two on top of each other? Please clarify.

Yes, the curve for [K]o+ = 7.21mM lies beneath the curve for [K]o+ = 5.21mM. This has been clarified in the figure caption.

(19) Text should note that the QIF iPRC does not develop a negative region at high K+ and high phase, as WB iPRC does.

We have updated the paragraph starting at line 409 to mention this.

(20) Pg. 16, line 4: “at the network scale” is cryptic - a more precise phrase would be preferable.

We have reformulated the sentence to clarify its meaning (see paragraph starting at line 380).

(21) Pg. 16, line 9: Reordering of words could make this clearer.

Done (see paragraph starting at line 385).

(22) Pg. 16: I am confused by line 14 because the big changes in the iPRC in Figure 9c do not align with the spike phase in Figure 9d. Please clarify what is meant here.

In the QIF model, at a given phase, the iPRC is the inverse of the slope of the voltage trace as a function of phase. Flatter slopes in Figure 9d therefore correspond to larger iPRC values in Figure 9c. This is illustrated in Figure S5. We have revised the paragraph starting at line 393 to make this point clearer.

(23) Pg. 16: Please clarify what is meant by a “delta synapse”.

By “delta synapse,” we meant a configuration in which each spike induces an instantaneous voltage jump in the postsynaptic neuron, modeled using the Dirac delta distribution. We have replaced the term “delta synapse” with “pulse-coupled,” which is more commonly used in the literature, and have added a clarification at its first occurrence in the manuscript.

(24) Discussion, line 2: Delete comma.

Done.

(25) Importantly, as noted above, the first par. needs to be rewritten since the presented method won’t work directly from data or from a target model for which most of the parameters, and hence the bifurcation diagram, are not known.

As mentioned above, we have included a new paragraph in the Discussion, starting at line 549, to clarify this point.

(26) Pg. 18: ”Originally” → ”Typically”, perhaps?

Done.

(27) Note the work of Marder et al. on temperature-related neural variability.

We thank the reviewer for this comment. We have added two relevant references from the work of Marder and colleagues addressing temperature-dependent neural variability and ionic concentrations in our manuscript (see the sentence starting on line 537).

(28) Pg. 20: Cut the ”Potassium dynamics and network models” subsection since it does not add anything substantive as written (or else expand it and include it in the subsection below).

We have expanded this paragraph and incorporated it into the subsequent subsection, as suggested by the reviewer.

(29) Finally, it’s a bit confusing that the authors refer to the potassium concentration as a slow variable yet have an instantaneous jump in this quantity at reset in their QIF model (i.e., instantaneous is VERY fast). Some explanation about this should be provided. Do they make the general assumption that ∆K is small, for example, such that this reset reflects the slow nature of K+ evolution (i.e., during the reset period, K+ would only change slowly, and hence by a small amount)?

Yes, ∆K is chosen to be small, to capture the behavior of the original model (compare for example panels c and h of Figure 6). As a result, in the QIF model, in the same way as in the original model, despite the fact that the dynamics of[K]o+ includes a fast component, on average [K]o+ evolves slowly. By using the averaging method, we can determine whether [K]o+ overall increases or decreases.

We clarified this in the manuscript, in the paragraph starting at line 243.

Reviewer #2 (Public review):

Summary:

The authors derive an integrate-and-fire model to describe the dynamics of a more complex Wang-Buzsaki model and compare the two models. A detailed discussion of bifurcation schemes in both models is convincing and allows us to evaluate the simpler model.

Strengths:

The idea is interesting, and the mathematical approach appears to be convincing. In addition, differences between the simple and original models are also discussed.

Weaknesses:

A comparison to experimental data is necessary to support the theoretical work.

As mentioned above in our answer to Reviewer 1, we demonstrate our method using as target the bifurcation diagram of a more realistic neuron model. Ideally, one would want to derive phenomenological models that capture bifurcation structures obtained from data. However, this is challenging and beyond the scope of the present study. We hope that this work encourages the development of such methods. We have revised the Introduction (see lines 52-57) and added a paragraph in the Discussion (see the paragraph starting at line 549) addressing this point.

Reviewer #2 (Recommendations for the authors):

The manuscript is well-structured; however, it appears that it has been edited with less care. Please see comments below:

(1) Page 2: “A third bifurcation, the saddle-homoclinic orbit (HOM) bifurcation,”: provide a reference for the bifurcation.

We have added a reference to the book by Izhikevich (see paragraph starting at line 63).

We have added additional references in the Introduction that we considered helpful.

(2) Page 3: “while a larger concentrations it is mediated by...”: remove “it”.

There was indeed a typo in this sentence. The intended phrasing is: “while at larger concentrations it is mediated by...”. We have corrected it accordingly (paragraph starting at line 134).

(3) Figure 2, caption: “dashed lines for unstable branches”: this is a dotted line.

Corrected to “dotted lines”. Thank you.

(4) Page 4: “is smaller than vSN (Fig. 3c),”: this figure panel does not exist, as well as the Fig.3d referred to afterwards. Please correct.

We intended to refer to Fig. 2. Figure references have been corrected. See paragraph starting at line 154.

(5) Page 6: “Fig. 3a-d shows ISN,0, vSN and a for [K]+o between 4 and 16 mM”: Fig.3c+d do not exist, please correct. Similar comment to “the absence of pump (Fig. 3e).” on the same page.

We intended to refer to Fig. 4. Figure references have been corrected (paragraphs starting at lines 199 and 209).

(6) Page 8: “(panel A)” → ”panel (a)”.

Done.

(7) Figure 4e: What is the meaning of the green dotted curve?

This curve represents the linear fit of the reset voltage vr from panel d of Fig. 4, to show that the minimal voltage values of the limit cycles are also well captured. We have added this curve to the legend and included a brief explanation in the figure caption.

Version 2

Abstract

In conductance-based models, spiking-induced ion concentration fluctuations can alter the excitability of single neurons. In particular, in class I models, variations in potassium concentration can induce qualitative changes in the dynamics through a codimension-2 bifurcation known as the saddle-node-loop (SNL). Investigating the implications of such effects at the level of neuronal networks will require computationally efficient single neuron models that still capture ion concentration dynamics realistically. To this end, we propose a method to derive a phenomenological model capturing the coupled extracellular potassium and voltage dynamics of a given class I conductance-based model. Rather than fitting voltage traces, we calibrate a canonical reduced model to the two-parameter bifurcation structure of the target model, with input current and a physiological parameter as coordinates. This preserves the location and type of dynamical transitions as the physiological parameter varies, allowing a single reduced model to capture neuronal dynamics across qualitatively different regimes. The resulting model is an extension of the quadratic integrate-and-fire model, in which extracellular potassium accumulation alters voltage dynamics by increasing the reset voltage. We apply our systematic reduction procedure to the Wang-Buzsáki model. Its phenomenological version exhibits quantitatively comparable dynamics and replicates the reshaping of the phase-response curve associated with the transition from SNIC to HOM spikes at elevated potassium. To illustrate the derived model’s applicability, we perform a preliminary investigation of how changes in potassium concentration influence synchronization in networks.

Introduction

The biophysical properties of neurons are well-established as critical determinants of both single-cell computation and network behavior. While complex neuron models allow a detailed analysis at the single-cell level, investigation of network dynamics often requires simpler models that can retain essential features of neuronal computation yet ensure computational efficiency [1–3]. This challenge is further intensified when biophysical single-cell properties change dynamically, such as through alterations in the local environment (e.g., temperature fluctuations, changing ionic concentrations) or neuromodulatory effects. To better understand how such modifications influence network dynamics, it is essential to incorporate these physiological changes into simplified neuron models as well. Here, we derive phenomenological neuron models of the integrate-and-fire type that quantitatively capture the dynamics of more complex conductance-based models not only for one specific set of the model’s parameters but also when a physiologically relevant parameter is allowed to vary.

In contrast to the common practice of fitting phenomenological models to spike trains or voltage traces recorded under a single physiological condition [4, 5], we propose to derive phenomenological models by matching a two-dimensional bifurcation structure spanned by the input current as well as an additional physiologically relevant parameter. Here, we demonstrate this approach using the bifurcation diagram of a biologically more complex model (specifically, a conductance-based model including potassium concentration dynamics) as the fitting target. In principle, the same strategy could be applied using a bifurcation diagram reconstructed from experimental data, provided that sufficiently rich measurements are available to characterize the relevant dynamical behavior. Our specific aim is to capture not only the firing rate, but also the bifurcation underlying the onset of regular firing at threshold, including switches in bifurcation type as the physiological parameter varies. While we focus here on the influence of extracellular potassium, the approach is equally applicable to other biological parameters.

In the processing of information within single cells, the dynamics of action-potential generation play an important role. Although influenced by numerous cellular characteristics, qualitative differences between regular action-potential firing (with all-or-none pulses at threshold) can arise from three distinct mechanisms. Their classification coincides but is not entirely identical with the excitability classes I and II described by Hodgkin (1948) [6] and observed earlier by Arvanitaki (1938) [7], which are based on the frequency of firing at spike onset (with arbitrarily low frequencies for class I and an immediate jump to higher frequencies for class II). These two classes are associated with distinct mathematical bifurcations that underlie the onset of regular spiking: the saddle-node-on-invariant circle (SNIC) bifurcation is the canonical class-I onset mechanism, while class II is typically associated with the subcritical Hopf bifurcation. Another bifurcation, the saddle-homoclinic orbit (HOM) bifurcation, shares the arbitrarily low onset frequencies of class I while differing in other properties (see Izhikevich [8] for details). The type of bifurcation is a critical determinant of the network state. For instance, neurons with SNIC dynamics tend to produce asynchronous activity when weakly coupled within inhibitory networks. In contrast, the transition to HOM dynamics in the same network configuration results in synchronous activity [9].

The transition from SNIC to HOM is of particular interest, as it is ubiquitous in the underlying mathematical framework (despite a quantitative dependence of the switching point on other parameters) and has the potential to induce dramatic changes in network behavior [10, 11]. One well-established parameter that causes this transition is the concentration of extracellular potassium. It is directly affected by neuronal spiking. While the outflow of K+ ions at each spike is typically compensated for by mechanisms such as Na+/K+-ATPases and glial buffering, these may not suffice to prevent transient changes during intense activity [12]. Such elevations alter the neuron’s excitability, forming a feedback loop that can create rich dynamics [13]. Moreover, abnormally elevated levels of extracellular potassium are involved in paroxysmal pathological activities that are specific to epilepsy (seizures) and migraine with aura (spreading depolarization) [14].

To include the effect of extracellular potassium concentration in simpler, network-compatible phenomenological models, we start from a conductance-based model including the potassium concentration dynamics and fit a simpler, integrate-and-fire-type model equipped with a potassium variable to this conductance-based model, aiming to capture both firing rates and the two-dimensional bifurcation structure with input current and extracellular potassium concentration as parametric dimensions. Fitting the bifurcation structure of the reduced model near the SNL bifurcation accurately captures the conductance-based model behavior under a range of ionic concentrations, associated with qualitative changes in excitability. This permits the study of transitions between those different regimes as potassium evolves dynamically. Accordingly, the phenomenological model captures a transition from SNIC to HOM firing with increases in extracellular potassium.

Overall, our contribution is twofold: a minimal model of potassium-dependent excitability and, more generally, a bifurcation-guided procedure for embedding physiologically meaningful parameter dependence into canonical reduced neuron models. In the following, we first demonstrate how to quantitatively derive the integrate-and-fire type version of a given class I conductance-based model. We then compare, at the single-neuron level, the dynamics of the two models and their phase-response curves. Finally, we investigate the effect of potassium on neuronal synchronization at the network level.

Results

Reducing a class I conductance-based model to a QIF neuron with dynamic [K+]o

We demonstrate our procedure on a Wang-Buzsáki model embedded in an extracellular space with dynamic potassium concentration (Fig. 1a) [13, 15], hereafter referred to as the target model. We reduce it to a model of type quadratic integrate-and-fire (QIF), which allows us to capture the normal forms of both SNIC and HOM bifurcations, as well as the transition between them at the so-called saddle-node-loop (SNL) point via the reset voltage (see Fig. S1). Reset values below the voltage at the saddle-node bifurcation of the QIF correspond to SNIC onsets, while larger values correspond to HOM onsets. We carefully choose parameter values of the QIF system and incorporate potassium dependencies to match the homoclinic, saddle-node, and SNIC bifurcation branches of the target model near the SNL transition. In particular, potassium concentration feeds back on the reset voltage, inducing the SNL. An additional reset rule, which increments potassium concentration by a small amount at each spike, mimics the flow of ions through voltage-gated potassium channels (Fig. 1b). Including an electrogenic Na+/K+-ATPase complements the potassium dynamics. The resulting reduced model is, like the original, a slow-fast system where slow changes in potassium drive the neuron across different excitability regimes.

Schematic representation of the two models: the target model (a) and its corresponding phenomenological version (b).

(a) Conductance-based model of excitability class I with dynamic extracellular potassium concentration, equipped with the usual Na+ and K+ voltage-gated channels and with an electrogenic Na+/K+ ATPase (diagram reproduced from [13]). (b) Quadratic integrate-and-fire (QIF) model with dynamic extracellular potassium concentration, equipped with an electrogenic Na+/K+ ATPase. Each spike (i.e., when the voltage v exceeds a threshold vth) triggers two reset rules: as in the classical QIF model the voltage is reset to a value vr, and the extracellular potassium concentration is additionally incremented by a small amount ΔK.

Original model

For convenience, we reproduce in “Methods” the definition of the extended Wang-Buzsáki model [13] (System (4)), which we use as an example to illustrate the reduction procedure.

Relying on the difference in timescale between [K+]o and the other variables, the fast subsystem of System (4) can be defined as Eqs. (4a) to (4c), where [K+]o is a parameter. At physiological potassium concentrations, spike onset is mediated by a SNIC bifurcation (e.g., when [K+]o = 5 mM, see Fig. 2a), while at larger concentrations it is mediated by a homoclinic bifurcation (e.g., when [K+]o = 12 mM, see Fig. 2b) [13]. Fig. 3a shows the SNL bifurcation organizing this transition.

Matching the bifurcation structure of the target model with the QIF model, at fixed values of [K+]o.

(a-b) Bifurcation diagrams of the fast subsystem of the Wang-Buzsáki model (Eqs. (4a) to (4c)) as a function of the applied current, featuring SNIC spike onset when [K+]o = 5 mM (a) and homoclinic onset when [K+]o = 12 mM (b). (c-d) Bifurcation diagrams of the model defined in Eqs. (1a) and (1b), with parameter values chosen to mimic the diagrams in (a) or (b), respectively. Limit cycles are shown in purple and fixed points in black, with solid lines for stable branches and dotted lines for unstable branches.

Bifurcation structure and frequency landscape near the SNL bifurcation.

(a) Two-parameter bifurcation diagram [K+]o versus applied current Iapp of the fast subsystem of the Wang-Buzsáki model (Eqs. (4a) to (4c)) and frequency of the limit cycles. (b) Same as (a), for the QIF model (Eqs. (1a) and (1b)). Note that in this model, the firing frequency goes to infinity as vr approaches vth. Here, we only show the frequencies up to a certain distance between those parameters.

When the full system slowly follows a branch of stable fixed points of the fast subsystem’s bifurcation diagram, i.e., when the neuron is quiescent, the dynamics of [K+]o can be studied using the standard slow subsystem (System 6). When the full system slowly follows a branch of stable limit cycles, i.e., when the neuron spikes, each spike causes a rapid increment of [K+]o (note the presence of v in Eq. (4d)). However, these increments are of small amplitude, and the dynamics of [K+]o remains slow on average (see already Fig. 6c). As is customary in such cases [16–18], to isolate the slow component of the dynamics, the averaging method can be applied to define the averaged slow subsystem (System 7). The strategy consists in averaging the derivative of the slow variable [K+]o over one period (i.e., one spike cycle), for each member of the family of fast subsystem stable limit cycles.

Step 1: QIF version of the Wang-Buzsáki model for a constant [K+]o

We consider the following QIF model: (1a) dvdt=a(v−vSN)2+c(Iapp−ISN,0−Ipump), (1b) ifv>vth, thenv←vr. Its branch of fixed points, defined by c(Iapp − ISN,0 − Ipump) = −a(v − vSN)2, folds at (Iapp = ISN,0 + Ipump, v = vSN). Stable limit cycles emerge via a SNIC bifurcation when the reset voltage vr is smaller than vSN (Fig. 2c), while they emerge via a homoclinic bifurcation at Iapp=ISN,0+Ipump−ac(vr−vSN)2 when vr > vSN (Fig. 2d), with a SNL bifurcation at vr = vSN. Let us first determine, for a given [K+]o, parameter values of System (1) such that its bifurcation diagram with respect to the applied current closely resembles that of the fast subsystem of the target model.

Note that in the target model (System (4)), Ipump depends only on potassium (Eq. (5d)). We can therefore treat this current as a potassium-dependent negative contribution to the applied current. In other words, when [K+]o is fixed, Ipump only horizontally translates bifurcation diagrams with respect to Iapp, such as those shown in Fig. 2. The parameter derivation below is performed in the absence of the pump current. Ipump can be added a posteriori, taken to be the same as in the target model (Eq. (5d)).

ISN,0 is the value of applied current Iapp at which the saddle-node bifurcation of interest occurs in the fast subsystem of the target model, in the absence of pump current. vSN is the membrane potential at the saddle-node. We take a and c to be the quadratic normal form coefficients for the saddle-node bifurcation. They are defined as (see, for example, [19]): a=12pTB(q,q),c=pT∂F∂Iapp, where p and q are the normalized left and right eigenvectors of the Jacobian matrix and B the Hessian quadratic form of the vector field F=(dvdt,dhdt,dndt)T at the saddle-node: Bi=∑j=1n∑k=1n∂2Fi∂xj∂xkqjqk Since the bias term c is close to 1 cm2 µF−1, we set it to this value for simplicity.

Then, if [K+]o is larger than the value at the SNL bifurcation in the fast subsystem of the target model [K+]o,SNL, we choose vr such that the homoclinic bifurcation in the QIF model takes place at the same value of applied current: (2) vr=−1a(Ihom,0−ISN,0)+vSN, where Ihom,0 is the value of Iapp at which the homoclinic bifurcation occurs in the fast subsystem of the target model, in the absence of pump current. Note that Ihom = Ihom,0 + Ipump is the value of Iapp at which the homoclinic bifurcation occurs in the presence of a pump current.

When [K+]o < [K+]o,SNL (i.e., in the SNIC case), we have more flexibility in the choice of the reset voltage. vr should be smaller than vSN to obtain a SNIC bifurcation. vr should also be reasonably close to the limit cycles’ minimum voltage in the target model. We choose to define vr in the SNIC case by extrapolating the linear fit of the values of vr assigned using Eq. (2) in the homoclinic case (i.e., when [K+]o > [K+]o,SNL) to lower values of [K+]o (see Fig. 4d). This satisfies the two criteria above.

Parameters of the QIF model (Eqs. (1a) and (1b)) depending on the potassium concentration.

(a-d) Derivation of the parameters ISN,0, vSN, a and vr. Note that the increase of the reset voltage vr with [K+]o reflects the increase of the potassium reversal potential EK in the conductance-based model (panel (d)). (e) Bifurcation diagram of the fast subsystem of the target model (Eqs. (4a) to (4c)) with respect to [K+]o, when Iapp = Iapp,SNL (i.e., the value of applied current at which the SNL bifurcation occurs) and Ipump = 0 µA cm−2. In this panel, we have reproduced the linear fit of the reset voltage vr shown in panel (d), for visual comparison with the minimal voltage values of the limit cycles.

Finally, the threshold voltage vth will allow us to approximate the limit cycles’ maximum voltage.

Step 2: QIF version of the Wang-Buzsáki model with dynamic [K+]o

We now compute parameter values of System (1) for a range of [K+]o, and depending on how strongly they vary with potassium, either set them to a constant or replace them with a function of [K+]o.

Fig. 4a-d shows ISN,0, vSN and a for [K+]o between 4 and 16 mM and vr only above [K+]o,SNL. We notice that vSN and the normal form coefficient a are relatively conserved in this interval. We fix them to their value at [K+]o,SNL. For ISN,0 and vr we use linear fits, constrained at [K+]o,SNL, that we extrapolate to lower concentrations in the case of vr. In conductance-based models, higher [K+]o weakens the potassium concentration gradient, thereby raising EK. The sodium and potassium reversal potentials typically bound voltage oscillations during spiking (see, for example, [20]), so the minimum voltage of spikes is expected to be higher when [K+]o is larger. This effect is reflected in the QIF model via the increase of the reset voltage vr (Fig. 4d).

For the threshold vth, we use a sigmoid that we extrapolate to lower concentrations, to fit the peak voltage of the limit cycles in the bifurcation diagram with respect to [K+]o, when Iapp = Iapp,SNL and in the absence of pump (Fig. 4e). This enables us to capture the shrinkage in spike amplitude that precedes the fold of limit cycles bifurcation. Note that, following the SNL bifurcation, the peak voltage increases slightly with [K+]o, by less than 1 mV. We do not include this increase in the QIF model.

We incorporate these functions of [K+]o in System (1), in which [K+]o now appears as a parameter. Its bifurcation structure with respect to [K+]o and the applied current Iapp closely resembles that of the fast subsystem of the original model (see already Fig. 3). A summary of the reduction procedure up to this stage is provided in the form of an algorithm in Table I. An illustration of this procedure using temperature instead of potassium concentration is shown in Fig. S2. Returning to the potassium example, we then include potassium dynamics to obtain the following model: (3a) dvdt=a(v−vSN)2+Iapp−ISN,0([K+]o)−Ipump([K+]o), (3b) d[K+]odt=−2γIpump([K+]o), (3c) ifv>vth([K+]o), thenv←vr([K+]o), (3d) K+]o←[K+]o+ΔK, with ISN,0([K+]o)=β0,ISN+β1,ISN[K+]o,vr([K+]o)=β0,vr+β1,vr[K+]ovth([K+]o)=α0,vth1+exp⁡(α1,vth([K+]o−α2,vth))−α3,vth

The evolution of the extracellular potassium concentration depends, on the one hand, on uptake by the Na+/K+-ATPase (Eq. (3b), where Ipump is given in Eq. (5d)), and on the other hand, on increments at each action potential (Eq. (3d)). Note that the reset rule Eq. (3d) captures the dependency of potassium dynamics on v in the original model (Eq. (4d)). When the target model is Wang-Buzsáki, the parameter values of the corresponding QIF version in the form of (3) are listed in Table II.

The fast subsystem of this model consists of Eqs. (3a) and (3c), with [K+]o treated as a parameter. Its bifurcation diagram with respect to input and potassium, and the frequency of the limit cycles, can easily be calculated analytically (see “Methods”). We can see in Fig. 3 that, with our reduction procedure, we achieve a very close match of the bifurcation branches emanating from the SNL bifurcation in the fast subsystem of the target model. In Wang-Buzsáki, the transition to depolarization block at larger potassium concentrations is mediated by a fold of limit cycles and a subcritical Hopf bifurcation (panel a). The dimensionality of the QIF fast subsystem does not allow for such bifurcations. Instead, we artificially obtain a depolarization block by silencing the dynamics when the threshold voltage vth drops to the reset level (see gray and green curves in Fig. 4e). Note that, although the firing frequency in the fast subsystem grows to infinity as [K+]o approaches this point, this does not effectively occur in the full system since [K+]o is incremented by a constant amount with each spike.

The standard slow subsystem of System (3), used to study the dynamics of [K+]o along stable branches of fixed points of the fast subsystem, is given by 0=a(v−vSN)2+Iapp−ISN,0([K+]o)−Ipump([K+]o)d[K+]odt=−2γIpump([K+]o). As in the original model, along stable branches of limit cycles of the fast subsystem, the dynamics of [K+]o includes a fast component, causing rapid increments at each spike (here, instantaneous jumps via the reset rule Eq. (3d)). The jump size ΔK is chosen to match the potassium increment per spike in the original model and is therefore small (see already Fig. 6h). Thus, the averaging method can still be employed (see, for example, [8]). The averaged slow subsystem of System (3) is then given by ⟨d[K+]odt⟩=ΔKT([K+]o)−2γIpump([K+]o) where T is the period of the limit cycles.

Our phenomenological neuron model closely captures the dynamics of the original conductance-based version

Under identical stimulation, the behavior of the phenomenological model resulting from this bifurcation diagram fitting procedure is very similar to that of the target model. In the same way as the original version, it can produce regular tonic firing (Fig. S4), as well as richer firing patterns. We illustrate this point with three scenarios typically occurring in class I conductance-based models [13, 16, 21], which can notably be obtained in the Wang-Buzsáki model (see overview in Fig. 5). They share the common feature of involving switches in firing regimes, caused by an activity-mediated accumulation of extracellular potassium. Those switches rely on: (i) the different regimes associated with each region of their common bifurcation diagram and (ii) slow (on average) changes in [K+]o. Although the time traces are not exactly the same, we will see that the two models are quantitatively comparable.

Three scenarios of activity-mediated switches in firing regimes that were reported in class I conductance-based models.

(a-c) Voltage traces, in the case of the Wang-Buzsáki model. (d) Schematic representation of trajectories in the [K+]o versus applied current bifurcation diagram through regions associated with various activity regimes.

Scenario 1: fold/homoclinic bursting

We first consider the hysteresis loop bursting mechanism described in [13], which relies on the interplay between slow changes in extracellular potassium concentration and voltage dynamics. It requires only fast-inactivating sodium channels, delayed rectifier potassium channels, and an indispensable electrogenic pump, without the need for any slow ion channels.

Starting from a physiological potassium concentration ([K+]o = 5 mM [22]), we stimulate the Wang-Buzsáki neuron with a constant deterministic applied current (Iapp = 0.25 µA cm−2). This corresponds to an initial condition in the spiking region (Fig. 6a), i.e., where the only attractor is a stable limit cycle with large voltage amplitude (Fig. S3a). There, potassium release through voltage-gated channels outweighs the effect of the pump on average over the spike cycles, leading to a progressive accumulation of extracellular potassium (first part of Fig. 6b).

Fold/homoclinic bursting scenario (deterministic input: Iapp = 0.25 µA cm−2).

For the Wang-Buzsáki model: (a) Bursting trajectory shown on the 2-parameter bifurcation diagram [K+]o versus Iapp of the fast subsystem. (b) Voltage and [K+]o time traces. (c) Enlargement of the burst highlighted in panel (b). (d) Bursting trajectory superimposed onto the 1-parameter bifurcation diagram of the fast subsystem with respect to [K+]o, zooming in on lower voltage values for better visibility. (e) Time derivative of the averaged and standard slow subsystems. For the quadratic integrate-and-fire model: (f-j) Same as (a-e).

When the system arrives in the area enlarged in Fig. 6a, where the bistable region is bounded from below (lower [K+]o) by the fold bifurcation and from above (higher [K+]o) by the homoclinic bifurcation, it begins to burst (second part of Fig. 6b). For completeness, note that this configuration is only a necessary condition for the bursting behavior to occur (for more details, see [13]). During each spiking phase, the system follows the branch of stable limit cycles (Fig. 6d), in the course of which potassium increases overall (in the averaged slow subsystem ⟨d[K+]odt⟩>0, see purple curve in Fig. 6e) because the increments at each action potential are not fully compensated by the pump (Fig. 6c). When it reaches the homoclinic bifurcation, the system switches to the branch of stable equilibria, initiating a quiescent phase. In the absence of spiking, the pump can restore potassium concentration down to the fold bifurcation (in the standard slow subsystem d[K+]odt<0, see the black curve in Fig. 6e), at which point the system jumps back to the stable limit cycle branch. This marks the start of a new bursting cycle.

Starting from the same initial voltage and potassium concentration, and stimulated with the same input, the integrate-and-fire version of the model successfully captures this bursting behavior (Fig. 6f-j). The underlying mechanism is the same as in the original model, with fold and homoclinic dynamical bifurcations governing the onset and termination of bursts. The small discrepancy between the two models in the number of spikes per burst and interspike interval is due to the divergence in the frequency of limit cycles (Fig. S3d) and slightly different location of the bifurcations in relation to potassium levels.

We should point out here that, as explained in Section 9.2.3 of [8], the reduction of the dynamics to the averaged slow subsystem is not valid in the close vicinity of the homoclinic bifurcation. Therefore, the intersection between the x-axis and the time derivative of the averaged slow subsystem (purple curve), visible in Fig. 6e and j, does not imply the existence of a stable limit cycle corresponding to tonic firing in the full system.

For this first scenario, in accordance with [13], we adopted a configuration where the potassium concentration rises until the neuron starts bursting. However, note that simulations can also be generated in which the neuron remains in the spiking region, for example, by implementing a buffering of the extracellular potassium. This can conveniently be done in both original and reduced versions of the model by introducing a term in the potassium equation. As a result, when the buffering is sufficiently strong, potassium stabilizes within the spiking region at a concentration corresponding to a fixed point of the averaged slow subsystem, which in this case equates to a limit cycle of the full system (Fig. S4).

Scenario 2: stochastic bursting

We now examine a switch in firing regime identified in a version of the Traub-Miles model [23] (a class I conductance-based model) accounting for potassium concentration dynamics. By introducing a noise component in the input, Contreras et al. obtained simulations showing that, after a period of tonic firing, the neuron switches to stochastic bursting, a regime where firing is intermittently interrupted [21]. It is possible to reproduce this behavior with the Wang-Buzsáki model.

We stimulate the neuron with colored noise (see Eq. (9) in “Methods”) of mean input Iapp = 0.3 µA cm−2, chosen sufficiently large not to interfere with the fold/homoclinic bursting behavior (Fig. 7a). Initially, like in the previous scenario, the neuron spikes periodically, causing an accumulation of extracellular potassium (first part of Fig. 7b). When the system reaches the bistable region, i.e., when [K+]o crosses the fold bifurcation of the fast subsystem, the firing pattern switches to stochastic bursting (second part of Fig. 7b). There, the noisy input triggers jumps between the two attractors: the stable fixed point corresponding to a resting state and the stable limit cycle corresponding to tonic firing (Fig. 7d).

Stochastic bursting scenario (noisy input, with mean Iapp = 0.3 µA cm−2).

For the Wang-Buzsáki model: (a) Trajectory shown on the 2-parameter bifurcation diagram [K+]o versus Iapp of the fast subsystem. (b) Voltage, [K+]o and applied current time traces. (c) Enlargement of the burst highlighted in panel (b). (d) Part of the trajectory corresponding to the burst highlighted in panels (b) and (f), superimposed onto the 1-parameter bifurcation diagram of the fast subsystem with respect to [K+]o. For the quadratic integrate-and-fire model: (e-h) Same as (a-d).

As expected from their common bifurcation structure, when confronted with the same stimulus (same realization of the stochastic process) as the target model, the integrate-and-fire version behaves very similarly (Fig. 7e-h).

Compared to the deterministic fold/homoclinic bursting, for which burst initiation and termination occur only at the boundaries of the bistable region, in this scenario the noise can induce switches between attractors within the bistable region (Fig. 7c-d,g-h).

On a side note, we observe an interruption of about 80 ms in the middle of the burst highlighted in Fig. 7c, during which the voltage is more depolarized than during interburst intervals. Instead, the system remains near the repelling slow manifold (the saddle branch in Fig. 7d). The transition from following an attracting manifold during the burst to evolving near a repulsive one is a canard-like effect, here induced by noise. Similar canard trajectories can be obtained in a deterministic way, see for example [24].

Scenario 3: transition to depolarization block

The last scenario we consider is the transition to depolarization block, which results in transient spike inactivation. This phenomenon is a pathological hallmark: it is the signature of cortical spreading depolarization at the single-neuron level and might also be involved in epileptiform activity [25]. The induction of a depolarization block, in the case where it is caused by an accumulation of extracellular potassium resulting from neuronal firing, has been extensively studied in conductance-based models with dynamic ion concentrations (e.g., [16, 21, 26–29]). We simulate it here with the Wang-Buzsáki model.

To achieve this, we use the same setup as for the stochastic bursting scenario, but without the noise component in the stimulus. As a result, in the bistable region, the system remains on the branch of stable limit cycles, allowing the extracellular potassium concentration to build up without interruption. When [K+]o exceeds the concentration corresponding to a fold of limit cycles of the fast subsystem, the neuron enters depolarization block (Fig. 8a-d). In comparison, in the stochastic bursting scenario, frequent pauses in firing due to the presence of noise make it easier for the regulatory mechanisms—in our model, the Na+/K+-ATPase—to maintain [K+]o in the bistable region, preventing the pathological block.

Depolarization block scenario (deterministic input: Iapp = 0.3 µA cm−2).

For the Wang-Buzsáki model: (a) Trajectory shown on the 2-parameter bifurcation diagram [K+]o versus Iapp of the fast subsystem. (b) Voltage and [K+]o time traces. (c) Enlargement of the transition to depolarization block shown in panel (b). (d) Part of the trajectory highlighted in panels (b) and (c), superimposed onto the 1-parameter bifurcation diagram of the fast subsystem with respect to [K+]o. For the quadratic integrate-and-fire model: (e-h) Same as (a-d).

When stimulated with the same deterministic input, the integrate-and-fire version of the model produces a voltage trace very similar to the depolarization block observed in the Wang-Buzsáki model (Fig. 8f-g). However, unlike the two other scenarios, this behavior does not entirely rely on the same mechanism as in the original model. The path leading to the block is unchanged, but the transition to the block itself cannot occur via the same dynamical bifurcation, due to the design of the reduced model. Instead, we emulate it by placing the system in a refractory state as soon as, at larger [K+]o values, the reset voltage meets the threshold value (Fig. 8e and h): the dynamics of both the voltage and [K+]o are stopped, and their values remain fixed thereafter (see Fig. 8f and g). With this constraint, although the dynamics during the block is only sketchily captured, the desired outcome is for the neuron to cease spiking. The dynamics preceding the block, by contrast, is captured satisfactorily: the QIF voltage trace features both the reduction in spike amplitude and increase in firing frequency (Fig. 8g). These two effects are in agreement with the experimental recordings in [27]. As shown in Fig. S3f and h, the frequency of the fast subsystem limit cycles as a function of [K+]o increases in two stages before the block. This pattern is a remnant of the situation with two homoclinic bifurcations at lower values of applied current (Fig. S3b and d).

Reshaping of the phase-response curve

We now compare the phase-response curves (PRCs) of the two models. For a spiking neuron, the PRC describes how the phase of the cycle at which a perturbation is received affects the timing of the next action potential. It provides a link between the dynamics of single neurons and their collective behavior (e.g., synchronization) when connected in a network.

Fig. 9a shows the infinitesimal PRC (iPRC, i.e., for infinitesimal input amplitude) of the Wang-Buzsáki model near spike onset for different potassium concentrations. Like when the SNL is reached via other biophysical parameters (e.g., see [30] for the capacitance), we can see a transition from almost symmetric at low [K+]o to left-tilted PRCs after the SNL bifurcation. Near spike onset, this alteration of the iPRC occurs suddenly at the SNL: the curves for [K+]o = [K+]o,SNL − 4 mM and [K+]o = [K+]o,SNL − 2 mM are still nearly identical.

Effect of [K+]o on the phase response curve and ability to synchronize near spike onset .

(a) iPRCs for the Wang-Buzsáki model (direct method). Note that the curve for [K+]o = 7.21 mM lies beneath the curve for [K+]o = 5.21 mM; the same holds for panel (c). (b) Spike voltage traces for the Wang-Buzsáki model. (c) iPRCs for the integrate-and-fire model (analytically, see “Methods”). (d) Spike voltage traces for the integrate-and-fire model. (e) Period of the limit cycles. (f) Locking range, assuming pulse coupling with a voltage perturbation of 0.15 µV. For the computation of the coupling function, pulses were normalized following Weerdmeester et al. [31].

The PRC reshaping at the SNL is well captured by the QIF model (Fig. 9c). In fact, this simplified model offers an intuitive understanding of how an increase in [K+]o deforms the iPRC. Since we are dealing here with a one-dimensional system, the iPRC is simply the inverse of the derivative of the spike voltage trace as a function of the phase. Flatter slopes therefore correspond to larger iPRC values. At low [K+]o, the inflection point of the spike is near phase (Fig. S5b), resulting in an almost symmetric iPRC (Fig. S5a). The effects of larger [K+]o are mainly expressed in the QIF model by a larger reset voltage, which results in skipping the beginning of the spike cycle (see Fig. 9d or compare panels b and d in Fig. S5). Increasing the reset is thus equivalent to retaining only a scaled version of the iPRC segment corresponding to late phases (see Fig. 9c or compare panels a and c in Fig. S5), resulting in more left-tilted iPRCs.

As pointed out in [30], the SNL bifurcation coincides with a reduction of the period at a given distance δIapp from spike onset (i.e., sharper f-I curves). This period reduction is captured by the QIF version of the model (Fig. 9e).

Note that, at large [K+]o, iPRCs contain a negative region (corresponding to phase delay) early in phase. This feature is not captured in the QIF model, related to the fact that QIF models do not explicitly describe membrane repolarization. Their iPRCs therefore cannot contain negative regions.

The changes associated with the SNL bifurcation directly affect the collective behavior of neurons in a network. Based on the theory of weakly coupled oscillators [32], and given a specific network configuration, the iPRC can be used to predict the neurons’ ability to synchronize. For example, in the case of two weakly pulse-coupled neurons (where each spike induces a small, instantaneous voltage jump in the postsynaptic neuron), the coupling function is a scaled version of the iPRC, and its odd part governs how the phase difference between the two neurons evolves over time [9, 31, 33]. When the uncoupled firing frequency differs between the two neurons, the locking range, i.e., the maximal frequency detuning such that there exists a stable fixed point for the dynamics of the phase difference, is twice the amplitude of the odd part of the coupling function [9]. Fig. 9f shows the locking range as a function of extracellular potassium concentration for a voltage perturbation amplitude of 0.15 µV. For both models, the locking range abruptly increases at the SNL, indicating enhanced synchronization ability. In more complex networks, the analysis must be adapted to the particular network architecture (e.g., network topology and synaptic coupling), but the PRC can still serve to predict aspects of the network state [34].

Extracellular potassium influences neuronal synchronization in a network

We now investigate the consequences of the changes introduced by the SNL bifurcation at the network scale. In particular, for tonically spiking neurons, we expect that the left-tilted PRCs at larger extracellular potassium concentrations induce in-phase synchronization of inhibitory networks [11]. We test this prediction in a basic network configuration. For this, we consider a network of 100 all-to-all pulse-coupled inhibitory neurons. We then impose on all these neurons a common extracellular potassium trace, simulating, for instance, an external wave reaching the network (Fig. 10a). With both models, the neurons of the network, which were previously out of synchronization (Fig. 10c and f), become in-phase synchronized upon arrival of this wave (Fig. 10d and g).

Response of a network of 100 neurons to an external potassium wave.

We considered an all-to-all connectivity scenario, with weak inhibitory pulse-coupling (δv = −0.000 15 mV). We used a noisy input (see “Methods”), common to all neurons, with mean Iapp = 0.25 µA cm−2. (a) Incoming potassium wave, common to all neurons. (b) 2-parameter bifurcation diagram [K+]o versus Iapp of the fast subsystem for the Wang-Buzsáki model. The crosses indicate the initial and final potassium concentrations. (c-d) Raster plot for 10 among the 100 neurons, during the third and 13th seconds, respectively. (e-g) Same as (b-d), for the QIF version of the model.

Disentangling the respective contributions of PRC asymmetry and frequency fluctuations to the effect of the SNL transition on synchronization requires further investigation. Furthermore, we deliberately chose [K+]o values such that the neurons remain in the tonic spiking region (Fig. 10b and e). Indeed, in the bistable region, where only a fraction of the neurons may be active, the PRC is not a reliable predictor of network behavior. A detailed study of the network dynamics is beyond the scope of this article (see, for example, [35]).

Discussion

Neuronal dynamics can influence network behavior, yet the inclusion of biophysical detail in larger network models both challenges computational efficiency and reduces model generality due to the number of additional parameters introduced. Commonly, neuron models are fitted to datasets from a single experimental setting, and one best-fit parameter set is determined. These models may fail to generalize when parameters not varied in the experiment are altered in the model. Here, we derive models that capture the impact of a biophysical parameter across a physiologically relevant range by fitting bifurcation diagrams rather than voltage traces tied to specific recording conditions. The resulting models capture the dynamical transitions across the fitted parameter range, at the cost of reduced precision at individual points in parameter space.

At the scale of a single neuron, we have shown that the phenomenological model accurately reproduces various activity patterns from the original model, without needing to adjust parameter values to each scenario. At the network level, preliminary simulations highlight the important role that extracellular potassium concentration plays in synchronization.

Specifically, we have here shown how to reduce a class I conductance-based model with dynamic potassium concentration to a quadratic integrate-and-fire model. Depending on the concentration of extracellular potassium and its dynamics, single-cell excitability in this model can vary between the SNIC and HOM types, capturing the regime transitions observed in the conductance-based neuron model. The effect of potassium on the bifurcation structure in the original, conductance-based model is in the reduced QIF model version implemented via a change in the reset voltage. Note that this approach is not limited to potassium; as illustrated for temperature in Fig. S2, it can be extended to other parameters known to induce an SNL transition.

A phenomenological neuron model capturing potassium dynamics

Phenomenological models that capture biophysical aspects are particularly valuable for exploring aspects at the network scale through systematic simulations. For example, phenomenological models of spike-frequency adaptation have been used to study network synchronization [2, 36]. In the same vein, our work formulates a simple model of potassium concentration dependence. It provides a minimal description of a dynamics that would typically require a more detailed, higher-dimensional system.

Extracellular potassium concentration is a particularly pertinent variable to retain in simplified neuron models. Typically, models assumed constant ion concentrations. This is motivated by the fact that the amount of relocated ions per action potential may have a negligible effect on the overall concentrations [37, 38], and that concentrations are regulated by mechanisms involving the Na+/K+-ATPase, cotransporters, and glial cells [39, 40]. Nonetheless, in the last decades numerous studies have focused on modeling their dynamics [13, 14, 16, 21, 27, 29, 41–47]. Indeed, experimental studies suggest activity-dependent fluctuations in concentration, in particular transient accumulations of extracellular potassium reaching several millimolars, in case of intense activity or following sensory stimulation [12, 39, 40, 45, 48, 49]. Such changes are expected to be particularly significant locally or in restricted spaces [39, 40]. Moreover, drastic increases in extracellular potassium are known to occur during pathophysiological events such as spreading depolarizations and epileptic seizures, as well as during hypoxia [39]. Both experimental evidence [21, 49–51] and results based on biophysical neuron models (e.g., [13, 14, 16, 21, 27, 29, 41, 47]) indicate that the weakening of the potassium driving force, resulting from activity-mediated extracellular build-up, in turn qualitatively alters neuronal excitability, at both moderate and large [K+]o elevations. This motivates the need for simple models capturing potassium dynamics to investigate the consequences of changes in neurons’ intrinsic excitability at the network scale.

There exist other minimalist models with potassium dynamics, mainly in the context of epilepsy [52–54]. The single neuron model by Depannemaecker et al. [52] reproduces a wide range of firing patterns, including seizure-like events, sustained ictal activity, and depolarization block, while being relatively simple. It remains, nevertheless, at a higher level of complexity compared to the present model, with its two-dimensional fast subsystem of type Hodgkin-Huxley and its two slow variables. The “epileptor-2” population model [53] by Chizhov et al. is also four-dimensional, with the advantage of explicitly modeling sodium dynamics. It can successfully simulate interictal and ictal discharges in high potassium conditions. However, the choice of coupling between potassium and voltage dynamics in this model does not permit onset bifurcation switches in individual neurons. In a recent study, Depannemaecker et al. propose an extension of the adaptive exponential integrate-and-fire model (AdEx) that accounts for impaired ionic gradients [54], by introducing a third, slow, variable.

Modeling the critical transitions in parameter space

The question of “what is a good model” has been debated extensively in the epistemological literature and many arguments in favor of simple leaky integrate-and-fire models have been presented [55]. For example, if well fitted, they have high predictive power for the voltage state variable. However, if one includes model predictions not only in observable state variables, but also for changed biophysical parameter sets, preserving the relevant bifurcation structure can improve robustness.

The idea that neuronal models should be understood through their bifurcation structure is not new in computational neuroscience. Beginning with FitzHugh’s phase-plane analysis and reduced models [56, 57], the emphasis gradually shifted from reproducing every ionic current to capturing the essential dynamical mechanisms underlying neuronal behavior. Building on this perspective, Rinzel and Ermentrout demonstrated that classes of neural excitability correspond directly to bifurcation types [58]. Rinzel also introduced slow–fast analysis to explain bursting dynamics [59]. Bertram et al. extended this viewpoint by classifying bursting according to the unfolding of codimension-2 or higher bifurcations of the fast subsystem [60], a framework that was subsequently broadened by Golubitsky et al. [61]. Izhikevich further organized this landscape by classifying bursters according to the pair of codimension-1 bifurcations that bound their silent and active phases [8, 62]. This framework continues to be developed and extended [30, 63, 64], making the construction of reduced models that match the relevant bifurcation diagram a natural next step.

Our work, therefore, presents a strategy to derive models that capture the dependency of the system on physiological quantities by matching the relevant bifurcation diagram. For a related approach in a broader context, see [65]. This way of constructing phenomenological models is a real alternative to approaches based on fitting spike trains, in particular because neurons are not stationary devices, but subject to constant change in their environmental parameters. Sources of variability include temperature [9, 66, 67], ionic concentrations [16, 21, 68] and neuromodulators [69]. These factors may directly depend on the recent history of activation [21, 70]. In some cases, their changes constitute a homeostatic response to other changes, helping to maintain the neurons’ properties in the face of a varying environment [71]. We argue that, when known, fitting the bifurcation diagram with respect to chosen parameters is a judicious way to obtain models tailored to the major source of instability in a given context, as we have done here with the extracellular potassium concentration. Fitting specific spike trains may result in models whose validity is largely restricted to the conditions of their construction. In contrast, fitting a bifurcation diagram may not only provide better robustness to changing conditions but also allow for the study of how these changes themselves slowly drive neuronal activity between different regimes.

In this study, we use as a target the bifurcation diagram of a conductance-based neuron model. This target is only indirectly constrained by experimental data, through gating functions fitted to voltage-clamp recordings. Ideally, the target bifurcation diagram would instead be obtained directly from experimental measurements. Estimating bifurcation diagrams from experiments, however, remains challenging. Some techniques exist to approximate branches of stable and unstable steady states, and branches of stable limit cycles can be approximated using slowly ramped current-clamp protocols [72]. In practice, however, many factors vary during recordings, and generating reliable diagrams would require tightly controlled experimental repetitions in various conditions, particularly for reconstructing two-dimensional bifurcation diagrams. Developing such protocols is beyond the scope of the present study but represents an important direction for future work.

There exist canonical phenomenological models that, thanks to their bifurcation landscape, can recapitulate the main features of neural excitability (see, for example, [73], or [63], which is based on the unfolding of a higher-dimensional bifurcation). However, they are formulated in terms of abstract variables. In some situations, having a model that contains specific bifurcations is not sufficient; their arrangement in terms of biophysical dimensions also matters. For example, the fold/hom bursting mechanism that we studied in the first scenario cannot occur when the unfolding of the SNL bifurcation is aligned with the [K+]o axis (see [13] for more details). Starting from a canonical model and adapting it to match desired bifurcation branches, as we did here with the QIF model, has the advantage of preserving important quantitative components.

Possible model extensions

Sodium dynamics

In this work, we restricted ourselves to potassium dynamics, although sodium concentrations are also directly affected by neuronal activity. This choice was made because potassium is the main player in activity-induced SNL transitions and switches in excitability class, while sodium is rather involved in spike amplitude reduction [21]. The engagement of the Na+/K+-ATPase to restore ionic gradients is accounted for in our model by the potassium dependence of the pump current. In [13], it is shown that sodium dynamics is not crucial for the bursting mechanism we studied in the first scenario.

Taking into account the dynamics of the two ion species, for example by tracking extracellular potassium and intracellular sodium concentrations, remains the most precise option [53, 74]. To avoid adding an extra variable, one could rely on electroneutrality to express one concentration as a function of the other [27, 28, 52]. Applying this strategy to our work is an interesting direction for refinement.

Potassium diffusion and spatial network models

A natural extension of the present model is its use in spatially structured networks in which extracellular potassium can diffuse between local compartments. The reduced model is particularly suited to this purpose because it retains the potassium-dependent transitions in neuronal excitability while remaining computationally inexpensive enough for large network simulations. At the single-neuron level, Fig. S4 already illustrates that adding a simple potassium diffusion term can stabilize extracellular potassium and thereby qualitatively alter the resulting firing regime. In a spatial network, allowing potassium released by active neurons to spread to neighboring extracellular compartments would introduce an additional, non-synaptic interaction between neurons, through which local activity can modify the excitability of nearby cells. Such models could be used to investigate how synaptic and potassium-mediated diffusive coupling interact to shape synchronization, spatial propagation of activity, and transitions between excitability regimes (see [75, 76]).

Transition to depolarization block

Finally, the way in which depolarization block is obtained in the reduced model (Scenario 3) is not ideal. Instead of being mediated by a supercritical Hopf bifurcation, or a subcritical Hopf coupled to a fold of limit cycles, as in conductance-based models, it was artificially achieved by silencing the dynamics when the reset and threshold voltages meet. Indeed, the fast subsystem of the reduced model, of type quadratic integrate-and-fire, cannot exhibit Hopf bifurcations. Instead, one would need to start from a two-dimensional fast subsystem, like the AdEx model. This would not only allow a more accurate modeling of the transition to the block but also open the possibility of studying the recovery from this pathological state.

Conclusion

In conclusion, we affirm the importance of phenomenological models that capture key underlying biophysical mechanisms. We contribute a model of potassium concentration dependence, readily suitable for integration into large networks. Moreover, our approach, based on the fitting of a bifurcation structure, provides a generalizable framework for deriving similar models for other relevant modulatory variables.

Methods

Extended Wang-Buzsáki model [13]

Model definition

The model obeys the system of differential equations (4a) Cdvdt=−INa−IK−IL−Ipump+Iapp, (4b) dhdt=5(αh(1−h)−βhh), (4c) dndt=5(αn(1−n)−βnn), (4d) d[K+]odt=γ(IK−2Ipump), with currents (5a) INa=gNam∞3h(v−ENa), (5b) IK=gKn4(v−EK), (5c) IL=gL(v−EL), (5d) Ipump=Ipump,max1+exp⁡(−k([K+]o−Keq)), and gating functions m∞=αmαm+βm,αm=0.1−v−35exp⁡(−35−v10)−1,βm=4exp⁡(−v−6018),αh=0.07exp⁡(−v−5820),βh=11+exp⁡(−v−2810),αn=0.01−v−34exp⁡(−v−3410)−1,βn=0.125exp⁡(−v−4480). The reversal potential for the potassium evolves with [K+]o according to EK=RTFlog⁡([K+]o[K+]i) where R = 8314.4 mJ K−1 mol−1 is the ideal gas constant, F = 96 484.6 C mol−1 the Faraday’s constant and T = 310 K the temperature. The variables and parameter values are listed in Tables III and IV.

Slow dynamics

The standard slow subsystem of System (4) is given by (6a) 0=−INa−IK−IL−Ipump+Iapp, (6b) 0=αh(1−h)−βhh, (6c) 0=αn(1−n)−βnn, (6d) d[K+]odt=γ(IK−2Ipump). As in [13], for [K+]o values such that there is a stable limit cycle in the fast subsystem, we define the averaged slow subsystem of System (4) as (7a) ⟨d[K+]odt⟩=γ(1T([K+]o)∫0T([K+]o)IK(v([K+]o;τ),n([K+]o;τ),[K+]o)dτ−2Ipump([K+]o)), where T is the period of the limit cycle.

Fast subsystem of the QIF model (Eqs. (3a) and (3c))

Frequency of the limit cycles

Let Iapp―= c(Iapp − ISN,0 − Ipump). If Iapp―>0 (tonic spiking region), the period of the limit cycles is given by: T=1aIapp―(arctan⁡(vth−vSNIappa)−arctan⁡(vr−vSNIappa)) If vr > vSN and −a(vr−vSN)2<Iapp―<0 (bistable region): T=log⁡(−2AB(C+1)+AB)2−aIapp― with A=(vr−vSN)−Iapp―a−Iapp―a,B=(vr−vSN)−Iapp―a+Iapp―a,C=vth−vSN−Iapp―a. Infinitesimal phase-response curve (iPRC)

Let vr―=vr−vSN. If Iapp―>0: iPRC(θ)=cos2⁡(Iapp―aTθ+arctan⁡(vr―Iappa))TIapp― If vr > vSN and −a(vr−vSN)2<Iapp―<0 : iPRC(θ)=−(−Iapp―acosh⁡(−Iapp―aTθ)−vr―sinh⁡(−Iapp―aTθ))2TIapp―(Iapp―a+vr―2)

Noisy input

To introduce noise in the applied current, we replace it with: (9) Inoisy=Iapp+η, where ηt is an Ornstein-Uhlenbeck (OU) process defined by dηt=−1τηηtdt+2σ2τηdWt Wt being the Wiener process. The OU process can be interpreted as a smoothly low-pass filtered white noise. In our simulations, we used a correlation time τη = 20 ms and a standard deviation σ = 0.05 µA cm−2.

Computational resources

To derive the parameters ISN,0, vSN, a and vr of the QIF model (1), we performed symbolic computations with the Python library Sympy [78]. In the case of vSN, we additionally used the numerical solver nsolve.

For numerical continuation, we used the software AUTO-07P [79].

Simulations of the neuron models were carried out with the Python package brian2 [80].